---
title: "Languages"
link-citations: true
csl: apa.csl
output:
  html_document:
    include:
      in_header: ./html/header_languages.html
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(DT)
```

This page includes a list of the idioms represented in the TALD database. The East Caucasian language family is very rich in terms of dialectal differentiation. There is no unified genealogical classification of all idioms spoken in Daghestan, so we compiled our own genealogical classification that tries to merge together different traditional and modern approaches to language/dialect classification of Daghestanian languages. In this tree you can also find the genealogical affiliation of each village represented in TALD. This classification is different from the one in [the Glottolog database](https://glottolog.org/resource/languoid/id/nakh1245). In the first tabset you can find a table that contains all the idioms represented in TALD and the chapter(s) in which they appear. In the second tabset you can find a tree that represents the genealogical classification that we use in the current version of TALD.

## {.tabset .tabset-fade .tabset-pills} 

### TALD Chapters by language

```{r}
list.files("data/orig_table", full.names = TRUE) |> 
  map(function(file){
    readxl::read_xlsx("data/contributors.xlsx", guess_max = 4000) |> 
    filter(render == 1) |> 
    rowwise() |> 
    mutate(id_0 = sprintf(str_c("%0", nchar(max(id))+1, "d_"), id),
           Chapter = str_c("- [",
                           title,
                           "](",
                           str_c(id_0, filename),
                           ".html)")) |> 
    select(filename, Chapter) ->
    features
  
  feature_dataset <- read_tsv(file, show_col_types = FALSE)
  
  feature_dataset |> 
    select(str_which(colnames(feature_dataset), 'feature\\d{1,}$')) |> 
    pivot_longer(cols = everything()) |> 
    distinct() |>  
    mutate(name = str_replace(name, 'feature', 'value')) ->
    columns_rename
  
  feature_dataset |> 
    select(language, idiom, matches("value\\d")) |> 
    select(-matches("numeric")) |> 
    rename_with(function(x){columns_rename$value[match(x, columns_rename$name)]}, matches('value\\d{1,}$')) |> 
    select(-matches("value\\d{1,}\\_name")) |> 
    mutate(across(everything(), as.character)) |> 
    pivot_longer(names_to = "Feature", values_to = "Value", -c(language:idiom)) |> 
    filter(!is.na(Value)) |> 
    rename(Language=language, 
           Idiom = idiom) |> 
    mutate(filename = str_remove(file, "data/orig_table/"),
           filename = str_remove(filename, ".tsv$")) |> 
    left_join(features, by = join_by(filename)) |> 
    select(Language, Idiom, Chapter)
}) |> 
  list_rbind() |> 
  mutate(across(everything(), as.factor))  |> 
  arrange(Idiom, Language) |> 
  distinct() |> 
  group_by(Idiom, Language) |> 
  summarise(Chapter = str_c(Chapter, collapse = "\n")) |> 
  mutate(Chapter = markdown::renderMarkdown(Chapter)) |> 
  arrange(Language) |> 
  DT::datatable(filter = 'top', 
                rownames = FALSE,
                escape = FALSE,
                options = list(
                  pageLength = 20, 
                  autoWidth = TRUE,
                  dom = 'fltpi'))
```

### Genealogical classification of TALD idioms 

```{r}
read_tsv("data/tald_villages.csv") |> 
  filter(is.na(standard) | standard %in% c("Standard Armenian")) |>
  mutate(village_dialect = str_c(village_dialect, " (", rus_village, ")")) |> 
  distinct(family, aff, default_level, dialect_toplevel, dialect_nt1, dialect_nt2, dialect_nt3, 
           village_dialect) |> 
  arrange(family, aff, default_level, dialect_toplevel, dialect_nt1, dialect_nt2, dialect_nt3, 
          village_dialect) |> 
  mutate(id = 1:n()) |> 
  pivot_longer(names_to = "cols", values_to = "value", -id) |> 
  na.omit() |> 
  group_by(id) |> 
  mutate(new_col = letters[1:n()]) |> 
  select(-cols) |> 
  ungroup() |> 
  pivot_wider(names_from = new_col, values_from = value) |> 
  mutate(L1 = a |> factor() |> as.double()) |> 
  group_by(a) |> 
  mutate(L3 = b |> factor() |> as.double()) |>
  group_by(b) |> 
  mutate(L5 = c |> factor() |> as.double()) |> 
  group_by(c) |> 
  mutate(L7 = d |> factor() |> as.double()) |> 
  group_by(d) |> 
  mutate(L9 = e |> factor() |> as.double()) |> 
  group_by(e) |> 
  mutate(L11 = f |> factor() |> as.double()) |> 
  ungroup() |> 
  group_by(f) |> 
  mutate(L13 = g |> factor() |> as.double()) |> 
  ungroup() |> 
  group_by(g) |> 
  mutate(L15 = h |> factor() |> as.double()) |> 
  ungroup() |> 
  pivot_longer(names_to = "col", values_to = "value", a:h) |> 
  filter(!is.na(value)) |> 
  mutate(L2 = ifelse(col == "a", "text", "children"),
         L4 = case_when(col == "b" ~ "text",
                        col %in% c("a") ~ NA,
                        TRUE ~ "children"),
         L3 = ifelse(is.na(L4), NA, L3),
         L6 = case_when(col == "c" ~ "text",
                        col %in% c("a", "b") ~ NA,
                        TRUE ~ "children"),
         L5 = ifelse(is.na(L6), NA, L5),
         L8 = case_when(col == "d" ~ "text",
                        col %in% c("a", "b", "c") ~ NA,
                        TRUE ~ "children"),
         L7 = ifelse(is.na(L8), NA, L7),
         L10 = case_when(col == "e" ~ "text",
                         col %in% c("a", "b", "c", "d") ~ NA,
                         TRUE ~ "children"),
         L9 = ifelse(is.na(L10), NA, L9),
         L12 = case_when(col == "f" ~ "text",
                         col %in% c("a", "b", "c", "d", "e") ~ NA,
                         TRUE ~ "children"),
         L11 = ifelse(is.na(L12), NA, L11),
         L14 = case_when(col == "g" ~ "text",
                         col %in% c("a", "b", "c", "d", "e", "f") ~ NA,
                         TRUE ~ "children"),
         L13 = ifelse(is.na(L14), NA, L13),
         L16 = case_when(col == "h" ~ "text",
                         col %in% c("a", "b", "c", "d", "e", "f", "g") ~ NA,
                         TRUE ~ "children"),
         L15 = ifelse(is.na(L16), NA, L15),
         L4 = ifelse(is.na(L3), NA, L4),
         L6 = ifelse(is.na(L5), NA, L6),
         L8 = ifelse(is.na(L7), NA, L8),
         L10 = ifelse(is.na(L9), NA, L10),
         L12 = ifelse(is.na(L11), NA, L12),
         L14 = ifelse(is.na(L13), NA, L14),
         L16 = ifelse(is.na(L15), NA, L16)) |> 
  filter(!is.na(value)) |>
  distinct(L1, L2, L3, L4, L5, L6, L7, L8, L9, L10, L11, L12, L13, L14, L15, L16, value) |> 
  mutate_all(as.character) |> 
  mutate(id = 1:n())  ->
  result
  
result |> 
  mutate(across(L1:L16, function(i){ifelse(i == "text", "icon", i)}),
         value = "FALSE") |> 
  bind_rows(result) |> 
  arrange(id) |> 
  select(-id) |> 
  rrapply::rrapply(how = "unmelt") |>
  rrapply::rrapply(condition = function(x) x == "FALSE",
          f = as.logical,
          how = "replace") |> 
  modify_tree(pre = function(x) {
    nm <- names(x)
    if (1 %in% nm) unname(x)
    else x
  }) |> 
  jsTreeR::jstree(search = TRUE, checkWithText = FALSE)
```
