---
title: "Team"
output:
  html_document:
    include:
      in_header: ./html/header_team.html
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

The project was designed by [Michael Daniel](https://orcid.org/0000-0002-2091-5713), [Konstantin Filatov](https://orcid.org/0000-0002-0706-9161), [Timur Maisak](https://orcid.org/0000-0001-6794-0978), [George Moroz](https://orcid.org/0000-0003-1990-6083), [Timofey Mukhin](https://orcid.org/0000-0002-4637-0627), [Chiara Naccarato](https://orcid.org/0000-0003-0017-6316), and [Samira Verhees](https://orcid.org/0000-0002-3479-7420).

[The TALD project](https://ilcl.hse.ru/en/projects/) would not have been realized in its present form without the advice and assistance of a number of people.

People who have contributed to the atlas in the past or are currently working on a topic include: Anastasia Alekseeva, Denis Arsentjev, Pavel Astafyev, Katerina Dagkou, [Michael Daniel](https://orcid.org/0000-0002-2091-5713), Timofey Dedov, [Nina Dobrushina](https://orcid.org/0000-0002-4559-4274), Maria Godunova, Fyodor Golosov, Anastasia Ivanova, Violetta Ivanova, Daria Izmalkova, Inga Kartozia, [Zaira Khalilova](https://orcid.org/0000-0003-1604-4510), [Evgeniya Klyagina](https://orcid.org/0000-0003-4720-8039), [Timur Maisak](https://orcid.org/0000-0001-6794-0978), Ekaterina Matyukhina, [Maksim Melenchenko](https://orcid.org/0000-0001-9040-2318), [George Moroz](https://orcid.org/0000-0003-1990-6083), [Timofey Mukhin](https://orcid.org/0000-0002-4637-0627), [Chiara Naccarato](https://orcid.org/0000-0003-0017-6316), Polina Nasledskova, Ivan Netkachev, Alexandra Nogina, Irina Politova, Rita Popova, Varvara Popova, Samira Ochs, [Manuel Padilla-Moyano](https://orcid.org/0000-0002-2051-2885), [Anastasia Panova](https://orcid.org/0000-0003-0793-671X), [Tatiana Philippova](https://orcid.org/0000-0002-9594-0833), [Ilya Sadakov](https://orcid.org/0000-0003-2338-7896), [Elena Shvedova](https://orcid.org/0000-0002-1823-1582), Pasha Skripkina, Daria Tsikh, Sara Zadykian, [Svetlana Zemicheva](https://orcid.org/0000-0003-3266-3696), Vasiliy Zerzele.

[Chiara Naccarato](https://orcid.org/0000-0003-0017-6316) coordinates the workshop for students.

[George Moroz](https://orcid.org/0000-0003-1990-6083) is responsible for the TALD website.

[Chiara Naccarato](https://orcid.org/0000-0003-0017-6316) and [Konstantin Filatov](https://orcid.org/0000-0002-0706-9161) are responsible for chapter editing.

[Elena Shvedova](https://orcid.org/0000-0002-1823-1582) is responsible for the TALD archive.

We are grateful to our colleagues Marina Chumakina, [Nina Dobrushina](https://orcid.org/0000-0002-4559-4274), Steven Kaye, [Yuri Koryakov](https://orcid.org/0000-0003-3962-3312), [Yury Lander](https://orcid.org/0000-0003-1798-7174), [Aigul Zakirova](https://orcid.org/0000-0003-4780-2732) for taking part in seminars and discussing with us some of the topics included in the Atlas.

Photos presented on the site are taken by [Michael Daniel](https://orcid.org/0000-0002-2091-5713), [Konstantin Filatov](https://orcid.org/0000-0002-0706-9161), [Timur Maisak](https://orcid.org/0000-0001-6794-0978), and [Timofey Mukhin](https://orcid.org/0000-0002-4637-0627).

## Contact us

You can always contact us via email at <cauatlas@gmail.com> or leave an issue on our [GitHub page](https://github.com/LingConLab/TALD).

## Our contributors

```{r}
readxl::read_xlsx("data/contributors.xlsx") |> 
  filter(render == 1) ->
  features

features$id_0 <- sprintf(paste0("%0", nchar(max(features$id))+1, "d_"), 
                         features$id)

features %>% 
  mutate(feature = ifelse(!is.na(major_topic_text), 
                          str_c("<a href='",
                                filename,
                                ".html'>",
                                title,
                                "</a>"),
                          str_c("<a href='",
                                str_c(id_0, filename),
                                ".html'>",
                                title,
                                "</a>")),
         author = str_split(author, ", ")) |>
  unnest_longer(author) |> 
  arrange(author, title) |> 
  group_by(author) |> 
  reframe(chapters = str_c(feature, collapse = "<br>")) |>
  mutate(last_name = str_extract(author, " .*?$"),
         last_name = str_squish(last_name)) |> 
  arrange(last_name) |> 
  select(-last_name) |> 
  DT::datatable(filter = 'top', 
                escape = FALSE,
                rownames = FALSE,
                options = list(pageLength = 150, 
                               autoWidth = TRUE,
                               dom = 'tpi'))
```

