---
title: "Team"
output:
  html_document:
    include:
      in_header: ./html/header_team.html
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

The project was designed by [Michael Daniel](https://orcid.org/0000-0002-2091-5713), [Konstantin Filatov](https://orcid.org/0000-0002-0706-9161), [Timur Maisak](https://orcid.org/0000-0001-6794-0978), [George Moroz](https://orcid.org/0000-0003-1990-6083), [Timofey Mukhin](https://orcid.org/0000-0002-4637-0627), [Chiara Naccarato](https://orcid.org/0000-0003-0017-6316), and [Samira Verhees](https://orcid.org/0000-0002-3479-7420).

TALD is a project of the [Linguistic Convergence Laboratory](https://ilcl.hse.ru/en/).

[Chiara Naccarato](https://orcid.org/0000-0003-0017-6316) and [George Moroz](https://orcid.org/0000-0003-1990-6083) coordinate the project and the workshop for students.

[George Moroz](https://orcid.org/0000-0003-1990-6083) and [Irina Politova](https://orcid.org/0009-0006-8265-7293) are responsible for the TALD website and the solving of technical issues.

[Chiara Naccarato](https://orcid.org/0000-0003-0017-6316), [George Moroz](https://orcid.org/0000-0003-1990-6083) and [Konstantin Filatov](https://orcid.org/0000-0002-0706-9161) are responsible for chapter editing.

[Elena Shvedova](https://orcid.org/0000-0002-1823-1582) is responsible for the TALD archive.

Photos presented on the site are taken by [Michael Daniel](https://orcid.org/0000-0002-2091-5713), [Konstantin Filatov](https://orcid.org/0000-0002-0706-9161), [Timur Maisak](https://orcid.org/0000-0001-6794-0978), [Timofey Mukhin](https://orcid.org/0000-0002-4637-0627), and [George Moroz](https://orcid.org/0000-0003-1990-6083).

## Contact us

You can always contact us via email at <cauatlas@gmail.com> or leave an issue on our [GitHub page](https://github.com/LingConLab/TALD).

## Our contributors

```{r}
readxl::read_xlsx("data/contributors.xlsx") |> 
  filter(render == 1) ->
  features

features$id_0 <- sprintf(paste0("%0", nchar(max(features$id))+1, "d_"), 
                         features$id)

features %>% 
  mutate(feature = ifelse(!is.na(major_topic_text), 
                          str_c("<a href='",
                                filename,
                                ".html'>",
                                title,
                                "</a>"),
                          str_c("<a href='",
                                str_c(id_0, filename),
                                ".html'>",
                                title,
                                "</a>")),
         author = str_split(author, ", ")) |>
  unnest_longer(author) |> 
  arrange(author, title) |> 
  group_by(author) |> 
  reframe(chapters = str_c(feature, collapse = "<br>")) |>
  mutate(last_name = str_extract(author, " .*?$"),
         last_name = str_squish(last_name)) |> 
  arrange(last_name) |> 
  select(-last_name) |> 
  rename(Author = author,
         Chapters = chapters) |> 
  DT::datatable(filter = 'top', 
                escape = FALSE,
                rownames = FALSE,
                options = list(pageLength = 150, 
                               autoWidth = TRUE,
                               dom = 'tpi'))
```

