```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = '')
library(lingglosses)
```

## Introduction

Comitative constructions express “accompaniment” relations, in which a “companion”, or secondary agent, accompanies the action(s) of the primary agent, or “accompanee”; cf. [@plungian2003: 170], [@stolzetal2006: 17]. In example (@one), Julie is the accompanee and Mark is the companion.

(@one) _Julie came **with Mark**_.

In the languages of the world, the companion in comitative constructions can be marked by means of different strategies, including case marking, clitics, and adpositions. Within the area of Daghestan, all such strategies are found. In the present chapter, we classify the languages of Daghestan according to **the strategy employed to mark the companion in comitative constructions**. By adopting a narrow definition of comitative construction [@arkhipov2009: 224-225], we exclude from the analysis complex constructions in which the predicate is repeated more than once (e.g. *John came and Mary came*, too) and coordinate constructions (e.g. *John and Mary came*).


## Results

In the languages of Daghestan, different strategies are employed to encode comitative meanings. In most languages, the companion is marked by means of case inflection (see [Case marking]). Other languages use clitics (see [Clitics]), or postpositions (see [Postpositions]).


### Case marking

The most common strategy to mark the comitative in the languages of Daghestan is case marking. Languages featuring this strategy can be further distinguished according to whether the comitative is marked by: a) a dedicated comitative case (see [Dedicated comitative case]); b) the instrumental case (see [Instrumental case]); or c) a spatial form (see [Spatial cases]).


### Dedicated comitative case

Dedicated comitative markers are found in four Lezgic languages (Agul, Archi, Rutul, Tsakhur). For Archi, Rutul and Tsakhur, the comitative case is reported to have instrumental semantics too [@mikailov1967: 55; @ibragimov2004: 57; @sosenskaja1999post: 122], but in these languages instrumental meanings are also marked by the ergative case. In Agul the comitative marker -*qaj* derives from the combination of a noun in the postessive case -*q* and a stative verb converb with a postessive prefix *q*-, so *dada-qaj* ‘with father’ would derive from `r gloss_example("＊dada-q qa-j", "father-POST POST.be-CVB", intext = TRUE)` ‘being at/near father’ [@maisak2014: 390].
In the other branches of the East Caucasian family, dedicated comitative markers are found in Standard Dargwa, Akusha, Kubachi, Mehweb, Itsari, Sanzhi, and Tanty (Dargwa),[^1] Tsova-Tush (Nakh),[^2] Lak and Khinalug. We also include in this group Northern Akhvakh, Karata and Tukita (Andic), whose comitative markers are sometimes described as postpositions. Given that they frequently attach to oblique forms of nouns/pronouns, we prefer to consider such markers as suffixes;[^3] cf. example (@two) from Karata.

(@two) Karata [@pasquereaudraft2020]
```{r}
gloss_example("den ɬo-**k’el**-ʕagi ʕurmi ge-da hač'e",
              "{I} who.OBL-**COM**-INDEF life do-IPFV COP.NEG",
              "I do not live with anyone.")
```


### Instrumental case

Some languages mark the companion by means of the instrumental case. We include in this group languages for which it is reported that the instrumental case has the main and most prototypical function of marking instruments, and comitative meanings are given as additional usages of the same case. These languages are Chechen and Ingush (Nakh), Kryz and Budukh (Lezgic), and Eastern Armenian; cf. example (@three) from Kryz.

(@three) Kryz [@authier2009: 214]
```{r}
gloss_example("riş gada-**zina** midfar-ik çi-u-ryu // ça-ba-rebe",
              "girl boy-**INS** wedding-SUB go-F-PRS.F {//} go-HPL-PRS.HPL",
              "The girl and the boy go to the wedding.")
```


### Spatial cases

Spatial cases mark the companion in all Tsezic languages, in which essive forms are used with different localization markers: cumessive in Bezhta, interessive in Khwarshi [@khalilova2009: 80], animate location essive in Hinuq [@forker2013: 98], apudessive in Tsez and Hunzib. Although in the literature the label “comitative” is sometimes used for such cases, it should be noted that they substantially differ from dedicated comitative markers in that they fit into the spatial inflection paradigm as any other spatial case. Spatial cases encoding comitative meanings can (but need not) be used in combination with postpositions meaning ‘with’ or ‘together’; cf. examples from Bezhta in (@four) and (@five).

(@four) Bezhta [@madieva1965: 76]
```{r}
gloss_example("kibːa-**ʁoj** ožo ø-eƛʼe-jo",
              "girl.OBL-**CUM** boy M-go-AOR",
              "The boy went away with the girl.")
```

(@five) Bezhta [@comrieetal2015: 202]
```{r}
gloss_example("jacːaː-**ʁoj** **q’ac’o** j-ec-al",
              "{(girl)friend}.PL.OBL-**CUM** **together** F-stay-INF",
              "Be with the girls.")
```

In Tabasaran (Lezgic) the companion is marked by the suffix -*q-di*, which is constituted by a localization marker meaning ‘behind’ (-*q*) and a comitative marker -*di*, which also conveys instrumental, manner and causer meanings when combined with other localization markers [@babaliyeva2013: 47-49]. In Udi (Lezgic) comitative meanings are encoded by the ablative case [@ganenkov2008: 39]. Finally, in Sanzhi Dargwa, the inelative suffix can be used as an alternative to the (more common) dedicated comitative case [@forker2020dargwa: 73, 572].


### Clitics

Clitics with comitative semantics are common in the Avar-Andic branch of East Caucasian languages. They are found in Avar, Andi, Bagvalal, Botlikh, Chamalal, Godoberi, and Tindi;[^4] cf. example (@six) from Bagvalal.

(@six) Bagvalal [@daraganmaisak2001: 177]
```{r}
gloss_example("wasː jaš**=eːna** w-aː",
              "brother girl**=COM** M-arrive.PST",
              "The brother arrived with a girl.")
```

A comitative clitic (=*žun*) is also found in Hunzib (Tsezic) as an alternative to case marking (see [Spatial cases]).


### Postpositions

Comitative postpositions are found in all Turkic languages in our sample, i.e. Azerbaijani, Kumyk, and Nogai; cf. example (@seven) from Standard Azerbaijani. These postpositions govern the genitive case with personal and demonstrative pronouns, and the absolutive case with nouns.

(@seven) Standard Azerbaijani [@tagievetal2006b: 540]
```{r}
gloss_example("mən-im**=lə** ged-ək",
              "{I}-GEN**=with** go-1PL(IMP)",
              "Come with me.")
```

Georgian features two postpositional constructions, one with the postposition =*urt* ‘with’ governing the instrumental case, and one with the postposition =*tan* ‘at’ governing the dative case and followed by the adverbial *ert-ad* ‘as one’; cf. (@eight).

(@eight) Georgian [@hewitt1995: 532]
```{r}
gloss_example("col-švil-it**=urt** // col-švil-Ø**=tan** **ert-ad**",
              "wife-child-INS**=with** {//} wife-child-DAT**=at** **one-ADV**",
              "with wife and child")
```

In Judeo-Tat the comitative is syncretic with the instrumental and is marked by a circumposition constituted by a locative proclitic *e*= and a dative enclitic =*(r)e* followed by the formant *voz* ‘with’; cf. example (@nine).

(@nine) Judeo-Tat [@authier2012: 109]
```{r}
gloss_example("**e=**tü**=revoz** **e=**biror-i**=revoz** zihis-de dan-üm",
              "**LOC=**you.SG**=INS** **LOC=**brother-MSD**=INS** live-INF (SUBJ)can-1SG",
              "May I live with you in brotherhood.")
```

Standard Lezgian (Lezgic) features the postposition *galaz* (i.e. the converb of the locative copula gala) governing the postessive case [@haspelmath1993: 225]; cf. example (@ten).

(@ten) Standard Lezgian [@haspelmath1993: 225]
```{r, results='asis'}
gloss_example("i kolkhoz-r-a lezgi-jr.i-qʰ **galaz** azerbajžan-r.i ermeni-jr.i stxawil.e-ldi zehmet č’ugwa-zwa",
              "DEM kolkhoz-PL-IN Lezgian-PL-POST **with** Azerbaijan-PL(ERG) Armenian-PL(ERG) fraternity-SUP.LAT work pull-IMPF",
              "Azerbaijanis and Armenians work fraternally together with the Lezgians in these collective farms.")
```

Postpositional constructions are also found in languages in which other strategies (i.e. case marking or clitics) are available, e.g. Avar, Botlikh, Godoberi, and Eastern Armenian.


## Distribution

The distribution of values on maps shows some areal and genealogical clustering. [Map 1](26_comitative_map.html#m1) shows the distribution of different types of marking, i.e. case marking, clitics and postpositions. Clitics are only found in Avar and in most Andic languages. Constructions with postpositions are found in all three Turkic languages in our sample, in Georgian, Judeo-Tat, and Standard Lezgian (Lezgic), in which comitative postpositions are described as deriving from converbs. Case marking is by far the most common strategy to encode comitative meanings in the languages of Daghestan. [Map 2](26_comitative_map.html#m2) shows the case form used to mark the companion in comitative constructions. A dedicated comitative case is found in four Lezgic languages, i.e. Agul, Archi, Rutul, and Tsakhur, all Dargwa languages in the sample, Tsova-Tush (Nakh), Lak, Khinalug, and the Andic languages Northern Akhvakh, Karata and Tukita.The instrumental case is employed in Chechen and Ingush (Nakh), Kryz and Budukh (Lezgic), and Eastern Armenian. Spatial cases are used in all Tsezic languages, Tabasaran and Udi (Lezgic).


[^1]: In Dargwa languages, the comitative case can also have instrumental semantics, but the same function is carried out by either a separate instrumental case, or the ergative.
[^2]: Comitative meanings in Tsova-Tush can also be conveyed by the instrumental (= ergative) -*w* [@desheriev1953: 262].
[^3]: For Karata it is reported that the suffix -*k’el* is obligatory added to oblique stems of personal pronouns, whereas nouns attach it to either the oblique stem or the absolutive stem [@pasquereaudraft2020].
[^4]: The comitative marker *k'ja(ː)* in Tindi is alternatively classified as a suffixal postposition governing the absolutive case [@magomedova2012: 58, 198] or as a particle [@magomedova2012: 46, 72]. According to Authier [-@authierdraft2020], this marker is probably a frozen converb.

## List of glosses {-}

```{r}
make_gloss_list(readxl::read_xlsx("../glosses.xlsx"))
```

## References {-}
