## Introduction

As in many other places, there is a custom in Daghestan to utter wishes when addressing someone who leaves for a journey. Formulae like “Have a nice trip” (English), “Bon voyage” (French ‘Good trip’), or “Sčastlivogo puti” (Russian, ‘Lucky way’) have parallels in the languages of Daghestan.

There is no single farewell formula typical for all Nakh-Dagestanian languages, but rather a variety of such formulae, for example (@bezhta1\-@kubachi3):

(@bezhta1) Bezhta (Mazhid Khalilov, p.c.)
```{r}
gloss_example(transliteration = "aaxo	 enƛ’e",
              glosses = "well 	M.go.IMP",
              free_translation = "Go well!")
```

(@tsova2) Tsova-Tush [@holiskygagua1994]
```{r}
gloss_example(transliteration = "dal-e-ⁿ			marš-ol		ḥoⁿ",
              glosses = "God-OBL-GEN	peaceful-ABSTR	2SG.DAT",
              free_translation = "God’s peace to you.")
```

(@kubachi3) Kubachi [@magometov1963: 205]
```{r}
gloss_example(transliteration = "salamatʰ 	w-aqʼ!",
              glosses = "good 		M-do.OPT",
              free_translation = "May you do well!")
```

Even within the same language community (such as a village), several farewell formulae can be used. For example, as found within a field study in Rutul villages, three different formulae can be used depending on the speaker and the village:

(@kufa4) Kufa Rutul (field data by Nina Dobrushina)
```{r}
gloss_example(transliteration = "hɨx-dɨ raˁq w-iši-j",
              glosses = "good-ATTR road N-become-OPT",
              free_translation = "‘Have a nice trip!’ (lit. ‘May the road be good’)")
```

(@kufa5) Kufa Rutul (field data by Nina Dobrushina)
```{r}
gloss_example(transliteration = "uɣur 		w-iši-j",
              glosses = "luck(AZ) 	N-become-OPT",
              free_translation = "‘Have a nice trip!’ (lit. ‘May the luck be’)")
```

(@kufa6) Kufa Rutul (field data by Nina Dobrushina)
```{r}
gloss_example(transliteration = "ačiχ-dɨ 		raˁq 	w-iši-j",
              glosses = "open-ATTR		road	N-become-OPT",
              free_translation = "‘Have a nice trip!’ (lit. ‘May the road be open’)")
```

Still, there usually exists one main formula which is the most frequent. In twelve Rutul villages which were surveyed within the same field research, the formula in (@kufa4) is by far the most common one. However, since most of the data analyzed in this chapter were collected from dictionaries, in most cases I did not have the opportunity to check which formula is the most common. Therefore, for each language I listed and mapped all formulae which were mentioned as expressing farewell meaning, and treated them equally.

## Results

There are two types of farewell formulae that cover large areas. The first type, shown on Map 1, literally means ‘May the road be good’ or ‘Good road!’.

(@mehweb7) Mehweb [@dobrushina2019: 141]
```{r}
gloss_example(transliteration = "hum-beʡ	aˤχ		d-uh-a-b!",
              glosses = "road-PL 	good 		NPL-become:PFV-IRR-OPT",
              free_translation = "May your road be good!")
```

(@lak8) Lak [@abdullaev2018]
```{r}
gloss_example(transliteration = "χːullu 		qʰin",
              glosses = "road		good",
              free_translation = "Good road!")
```

(@lezgian9) Lezgian (dialect of Tsnal) (Fazir Dzhaliev, p.c.)
```{r}
gloss_example(transliteration = "qsin	req	xu-j",
              glosses = "good	road	become-OPT",
              free_translation = "May the road be good!")
```

This formula seems trivial; many European languages use similar wishes (see the French and English examples above). But, as will be shown later, in Daghestan it shows an interesting distribution.

The second type, presented on Map 2, literally means ‘May the road be straight’:

(@andi10) Andi (dialect of Zilo) (Aigul Zakirova, p.c.)
```{r}
gloss_example(transliteration = "miq’i		b-it’i-ɬi-do",
              glosses = "road		N-be.straight-INCH-OPT",
              free_translation = "May the road be good!")
```

(@avar11) Avar [@gimbatov2006]
```{r}
gloss_example(transliteration = "nuχ		bit’a-gi!",
              glosses = "road		be.straight-OPT",
              free_translation = "May the road be straight!")
```

(@hunzib12) Hunzib [@vandenberg1995: 88]
```{r}
gloss_example(transliteration = "huni 	bic’i		r-aq-ob",
              glosses = "road 	straight		V-happen-OPT",
              free_translation = "May the road be straight!")
```

This formula is found almost exclusively in Avar-Ando-Tsezic languages. Vice versa, there  is only one language of this group for which I found no traces of such a formula in the grammar and dictionary — Hinuq. In all Avar-Ando-Tsezic languages the formula has a very similar lexical structure, and the word ‘straight’ is cognate.

An interesting case is found in Archi, a Lezgic language surrounded by speakers of Avar and Lak. The Archi farewell formula contains a word which is not morphologically transparent, and speakers cannot translate it, but it clearly contains the root *t’al-*, which means ‘straight, correct’. Archi was mapped together with the languages which have formulae meaning ‘may the road be straight’, although its formula is not fully identical with them.

(@archi13) Archi (Bulbul Musaeva, p.c.)
```{r}
gloss_example(transliteration = "t’al-ek-mul 		ka",
              glosses = "straight-?-NMLZ 	become.IMP",
              free_translation = "Have a nice trip!")
```

There is one more pattern which covers several languages. In most Lezgic languages and Khinalug there is a formula containing a word which originates from Azerbaijani, where it means ‘luck’ and is used in different wish formulae, but is not common at parting. By contrast, at least in some Lezgic languages, this word is used exclusively in farewell formulae. For example, in the dictionary of Rutul the word uʁur (*угъур*, [@alisultanovsulejmanova2019]) has no translation of its own, and all examples of its usage are farewell wishes. In a similar way, my Lezgian consultant was unable to translate this word and was unaware of its origin.

(@Agul14) Agul (Huppuq’ dialect) (Solmaz Merdanova, p.c.)
```{r}
gloss_example(transliteration = "jaʁur 		xu-raj",
              glosses = "luck(AZ) 	be-OPT",
              free_translation = "Have a nice trip!")
```

(@lezgian15) Lezgian  [@talibovgadzhiev1966: 397]
```{r}
gloss_example(transliteration = "juʁur		xuj",
              glosses = "luck(AZ)	become.OPT",
              free_translation = "Have a nice trip!")
```

The form of this word slightly varies in different Lezgic languages (*jaʁur* in Agul, *juʁur* in Standard Lezgian, *uɣur* and *uʁur* in different dialects of Rutul), but the formula is identical in all these languages: luck[AZ] N-become-OPT. Interestingly, a parallel formula *uğur-lu ol-sun* (luck-ADJ be-JUSS) exists in modern Azerbaijani, but is not dedicated to the farewell wish, and can be translated as ‘good luck!’. The most common farewell formula in Azerbaijani is *yaxşı yol* ‘good road’; in the above mentioned field research in Rutul villages this Azerbaijani phrase came up once, but in general it is not used in send-offs in Rutul villages.

Apart from these three types, there are other, less frequent formulae which are not mapped. For example, in the adjacent language communities of Tsez and Hinuq the same formula, not found in any other places, is used. It consists of the words ‘road’ and ‘send / let’, but the meaning is not transparent; most likely, the intended but not overtly expressed subject of this formula is God: ‘May [God] let the road’ (‘пусть Бог отпустит дорогу’, as it was translated by the speaker of Tsez Arsen Abdullaev).

(@tsez16) Tsez  [@xalilov1999: 109]
```{r}
gloss_example(transliteration = "huni 	r-egir-o-ƛ",
              glosses = "road 	IV-send/let-IMP-OPT",
              free_translation = "Have a nice trip! (lit. ‘May road let’)")
```

(@hinuq17) Hinuq  [@forker2013: 231]
```{r}
gloss_example(transliteration = "hune 	kekir-ƛo",
              glosses = "road 	send/let-OPT",
              free_translation = "Have a nice trip!  (lit. ‘May road let’)")
```


## Distribution

Similarly to what was found for morning greetings ([@naccaratoverhees2021]), most farewell formulae show a geographical distribution roughly corresponding to the areas where two large lingua francas of Daghestan operated - Avar (in Central Daghestan) and Azerbaijani (in South Daghestan).

The first area is covered by the formula ‘May the road be straight’. All Avar-Ando-Tsezic languages (apart from, probably, Hinuq) have it. As mentioned above, there is also one language of another branch whose farewell formula most likely developed under the influence of Avar — the Lezgic language Archi.

The second area features the formula with the Azerbaijani word ‘luck’; it is found in all Lezgic languages of Daghestan apart from Archi (which indicates the areal spread of this formula), and in Khinalug.

The most widely spread formula, ‘May the road be good!’ (or ‘Good road!’), can not be attributed to contact influence with any degree of confidence, since it is found all across the world. However, in my sample  it did not come up in any Avar-Ando-Tsezic languages, probably because of the highly influential formula ‘May the road be straigh
t’. Otherwise its area cannot be well defined, since it is found in Chechen, Rutul, Lak, Kumyk and other geographically and genetically dispersed languages.

The two maps reflect the spread of these three formulae. The first map compares the distribution of the formula containing the words ‘straight road’ to the formula containing the words ‘good road’. The second map compares the distribution of the ‘straight road’ formula to that with the Azerbaijani word ‘luck’.

## List of glosses {-}

```{r}
make_gloss_list(readxl::read_xlsx("../glosses.xlsx"))
```

## References {-}
