```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = '')
library(lingglosses)
```

## Introduction

The chapter deals with first person plural pronouns, i.e. pronouns expressing the meaning ‘we’. Languages of the world differ as to whether they distinguish between pronouns with exclusive and inclusive reference. The languages that make the distinction have two pronouns in the 1PL: ‘we, you are included’ and ‘we, you are excluded’. The Khnov dialect of Rutul (Lezgic < East Caucasian) is an example of such a language. Example ([1a](#one)) shows the context ‘You haven’t seen us’, where the hearer is not included in the reference of ‘us’, and the form *juˤqumbɨ* is used. In ([1b](#one)) another form is used: *jinewi*; in this case the hearer is included in the reference ‘We are all humans’. The form *juˤqumbɨ* is called “first person exclusive” (EXCL), and *jinewi* “first person inclusive” (INCL).

(@one) Khnov Rutul, Lezgic, East Caucasian [Ilya Sadakov, field data]
  a.
```{r}
gloss_example(transliteration = "wa-s juˤqumbɨ ʁ-a<d>gu-d dɨš",
              glosses = "2SG-DAT EXCL PV-<HPL>see.PFV-ATTR COP.NEG",
              free_translation = "You haven’t seen us.")
```
  b.
```{r}
gloss_example(transliteration = "jinewi sine insan-ar d-iʔi",
              glosses = "INCL all person-PL HPL-COP",
              free_translation = "We are all humans.")
```

In East Caucasian languages the inclusive and the exclusive may or may not be differentiated. In the latter case there is only one form which expresses the meaning ‘we’, both with exclusive and inclusive reference. This is the case in Kina Rutul (the variety of Rutul spoken in the village of Kina). In both contexts ([2a](#two)) (the hearer is excluded) and ([2b](#two)) (the hearer is included) the form *je* is used, which contrasts with Khnov Rutul (see ([1a](#one)) and ([1b](#one)) above).

(@two) Kina Rutul, Lezgic, East Caucasian [Maxim Melenchenko, field data]
  a.
```{r}
gloss_example(transliteration = "wa-s je ʁ-e<d>gü-r=a dɨš",
              glosses = "2SG-DAT 1PL PV-<HPL>see.PFV-CVB=be COP.NEG",
              free_translation = "You haven’t seen us.")
```
  b.
```{r}
gloss_example(transliteration = "je sin insan-ar d-iʔi",
              glosses = "1PL all person-PL HPL-COP",
              free_translation = "We are all humans.")
```

Cross-linguistically, the formal distinction between inclusive and exclusive pronouns is quite common. It is present in 68 out of 200 languages (34%) according to WALS data [@cysouw2013]. In this chapter and in the map I will use the label “attested” if the inclusive/exclusive distinction is present in a language, and “not attested” if it is not present. In the latter case, I will designate the first person plural pronoun as **neutral** ‘we’.

The goal of this chapter is to classify the languages of Daghestan according to the presence of the inclusive/exclusive distinction in their pronominal systems.


## Results

### Non-East-Caucasian languages
None of the non-East-Caucasian languages included in the sample distinguishes between the inclusive and the exclusive. These are the following languages: Armenian, Georgian, Tat (Muslim Tat and Judeo-Tat), Kumyk, Nogai and Azerbaijani. The forms of speech act participants (SAP) pronouns of these languages are shown in [Table 1](#t1).

#### **Table 1.** SAP pronouns in Armenian, Georgian, Tat, Kumyk, Nogai and Azerbaijani (see [@dum-tragut2009: 124 ; @hewitt1995: 76 ; @griunberg1963: 38 ; @abdullaevaetal2014: 247 ; @baskakov1973: 161 ; @shiralievsevortyan1971: 79], respectively). {#t1 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lk6{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1PL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Modern East Armenian</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">jes</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">menkʰ</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">du</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">dukʰ</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Standard Georgian</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">me</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">čven</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">šen</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">tkven</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Muslim Tat</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">mæn</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">(i)mu(n)</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ty</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">(i)šmu(n)</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Standard Kumyk</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">men</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">biz</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">sen</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">siz</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Standard Nogai</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">men</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">biz</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">sen</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">siz</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Standard Azerbaijani</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">mən</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">biz</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">sən</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">siz</span></td>
  </tr>
</tbody></table>

### Avar-Andic branch

In the Andic branch of East Caucasian, the inclusive/exclusive distinction was attested in all languages. The pronouns of Karata of Archo ([Table 2](#t2)) serve as an example of a system typically found in Andic languages.

#### **Table 2.** SAP pronouns of the Karata language, Archo village [@magomedbekova1971: 101]. {#t2 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-reig{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:left;vertical-align:top}
.tg .tg-7lk6{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2">INCL</th>
    <th class="tg-reig">EXCL</th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2">Karata, Archo village</td>
    <td class="tg-7lk6">den</td>
    <td class="tg-7lk6">iƛi</td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">išːi</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">men</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">bišːdi</span></td>
  </tr>
</tbody></table>



In Avar, almost all the dialects and lects also have the distinction. Only two varieties lack it. According to [@isakov2020: 97-98] the idioms of the Kusur and Upper Dzhengutaj villages do not distinguish between inclusive and exclusive. In the case of Kusur, this can be due to its geographical isolation from other Avar villages and to the fact that Kusur is adjacent to the territory of Tsakhur (Lezgic), which lacks the inclusive/exclusive distinction. [Table 3](#t3) compares the pronoun sets in Standard Avar and in the Avar of Kusur. We see that the latter lost the inclusive form *niɬ*, and the meaning of the exclusive pronoun *niž* expanded to cover the neutral ‘we’.

#### **Table 3.** Standard Avar pronouns [@alekseevataev1997: 53-54] and the pronouns of the Avar of Kusur [@isakov2020: 97-98]. {#t3 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lk6{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">INCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">EXCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Standard Avar</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">dun</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">niɬ</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">niž</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">mun</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuž</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Kusur Avar</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">dun</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">niž</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">mun</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuž</span></td>
  </tr>
</tbody></table>

### Dargwa languages {#dargwa -}

Dargwa languages (here I use the classification of Dargwa languages proposed by Yury Koryakov [-@koryakov2021]) demonstrate a great variation with respect to the inclusive/exclusive distinction. All Northern Dargwa languages lack the distinction. Some but not all Southern Dargwa languages as well as Chirag and Kaitag Dargwa make the distinction. Dargwa languages are remarkable because of the diversity of the diachronic processes in personal pronouns systems of different varieties.

Ganenkov [-@ganenkov2013] reconstructs the pronominal system of Proto-Dargwa and proposes a scenario of its evolution in languages which descended from the proto-language. [Table 4](#t4) shows the reconstruction and the forms attested in some Dargwa varieties.

#### **Table 4.** Personal pronouns in Proto-Dargwa, Chirag, Kaitag, Northern Dargwa — Standard Dargwa, Tsudakhar and Butri, Southern Dargwa — Shiri, Kunki, Itsari, Sanzhi, Tanty and Kubachi. {#t4 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-vt8r{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lk6{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">source</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">INCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">EXCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Proto-Dargwa</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@ganenkov2013</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">*du</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">*nux:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">*nus:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">*ħu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">*nuš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Chirag</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@kibrikkodzasov1990: 221; Dmitry Ganenkov, p.c.</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">du</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">núxːa</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">núsːa</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕuˤ</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">núš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Kaitag</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@temirbulatova2004: 130</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">du</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nixːa</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nisːa</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">i</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Standard Dargwa</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@vandenberg2001: 27</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nu </span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">(GEN</span><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent"> dila</span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">)</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuša</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ħu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ħuša</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Tsudakhar</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@abdullaev1954: 141</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">du</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nux:a </span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">(OBL </span><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">niš:-</span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">)</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕuš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Butri</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@shakhbanova2007: 25, 91, 101-102</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">du</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nux:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">i</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">iš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Shiri</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@belyaev2019: 34; Oleg Belyaev, p.c.</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">du</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuxːa</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nusːa</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʔü</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Kunki</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">Dmitry Ganenkov, p.c.</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">du</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuš:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nusːa</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">u</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Itsari</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@sumbatovamutalov2003: 37</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">du</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuš:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">u</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Sanzhi</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@forker2020dargwa: 20</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">du</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuš:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">u</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Tanty</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@sumbatovalander2014: 77</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">du</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuxːa </span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">(OBL </span><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">niš:-</span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">)</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕuˤ</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕuˤx:a </span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">(OBL </span><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕaˤš:-</span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">)</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Kubachi</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@magometov1963: 139</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">du</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nus:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">u</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uš:a</span></td>
  </tr>
</tbody></table>

According to Ganenkov’s reconstruction, in Chirag, Kaitag and Shiri, the Proto-Dargwa pronominal system remained unchanged. In other varieties, except Kunki, the inclusive/exclusive distinction was lost. In Kunki the distinction is still present, but according to the hypothesis, the original system underwent a change. The original \*INCL pronoun \**nux:a* was replaced by the descendent of the original \*2PL \**nuš:a*. The latter was substituted, in its 2PL function, by an innovative 2PL *uš:a*. This new 2PL form appeared in all varieties except Chirag, Kaitag and Shiri, and also in Itsari, where the 1PL and the 2PL are syncretic. Ganenkov suggests that the innovative form *uš:a* (and its parallels in other innovating systems) has been formed from the 2SG pronoun by analogy.

As for the varieties where the distinction was lost, three sources for their present 1PL forms are attested: \*INCL, \*EXCL and \*2PL. In Butri the source of the 1PL is the former inclusive pronoun. A similar process occurred in Tsudakhar and Tanty, but there the former inclusive took up the function of 1PL only in the absolutive and ergative case, while in other cases the stem *niš:*- which originates from \*2PL is used. In Kubachi, the 1PL goes back to the \*EXCL. In the remaining cases, i.e. in Standard Dargwa, Itsari and Sanzhi, the \*2PL became the 1PL, and the original \*INCL and \*EXCL were both lost. In Itsari, there is a syncretism of 1PL and 2PL, whereas in Standard Dargwa and Sanzhi the \*2PL shifted to the 1PL, and a new form (similar to the situation in other varieties) for the 2PL appeared. A notable change occurred in Standard Dargwa and some Northern dialects, where the form of 1SG (originally, *du*) was replaced by *nu*, so that the system became symmetrical: *nu* — 1SG, *ħu* — 2SG, and the plural forms are “derived” from the singular ones by the means of formant -*ša*.

[Table 5](#t5) summarizes the origins of pronouns in each idiom. In the columns “1SG”, “INCL”, etc., the origin of the corresponding present pronominal form is shown. E.g., if the column “INCL” contains \*2PL, this means that the synchronic form of INCL originates from \*2PL of proto-Dargwa. If no origin is displayed, the synchronic form is written, e.g. *ħuša* for Standard Dargwa, which does not have a correspondence in the proto-language and is an innovation.

#### **Table 5.** The origins of personal pronouns in the Dargwa varieties, according to [@ganenkov2013]. {#t5 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-vt8r{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lk6{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">INCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">EXCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Proto-Dargwa</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">*du</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">*nux:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">*nus:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">*ħu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">*nuš:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">*nuš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Chirag</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*1SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*INCL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*EXCL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2PL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">núš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Kaitag</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*1SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*INCL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*EXCL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2PL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Standard Dargwa</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nu </span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">(GEN</span><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent"> </span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*1SG)</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2PL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ħuša</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ħuša</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Tsudakhar</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*1SG</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">\*INCL</span><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent"> </span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">(OBL \*2PL)</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕuš:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕuš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Butri</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*1SG</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*INCL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">iš:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">iš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Shiri</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*1SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*INCL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*EXCL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2PL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Kunki</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*1SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2PL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*EXCL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uš:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Itsari</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*1SG</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2PL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2PL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nuš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Sanzhi</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*1SG</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2PL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uš:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uš:a</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Tanty</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*1SG</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">\*INCL</span><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent"> </span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">(OBL \*2PL)</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕuˤx:a </span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">(OBL </span><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕaˤš:-</span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">)</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕuˤx:a </span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">(OBL </span><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʕaˤš:-</span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">)</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Kubachi</span></td>
    <td class="tg-vt8r"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*1SG</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*EXCL</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">*2SG</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uš:a</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uš:a</span></td>
  </tr>
</tbody></table>

Dargwa languages show multiple semantic shifts in the system of personal pronouns.

### Lezgic languages

Lezgic languages also show diverse behavior with respect to the inclusive/exclusive distinction. Some languages have it, including Tabasaran, Agul and Archi, while some others lack it, including Budukh, Tsakhur, Udi and Lezgian. In this section, we consider pronominal systems of all Lezgic languages except Archi. Archi and Lak, whose systems are both very peculiar and very similar, are discussed further below in a separate section.

[Table 6](#t6) shows the personal pronouns in Tabasaran and Lezgian.

#### **Table 6.** Personal pronouns in Tabasaran [@babaliyeva2013: 54] and Lezgian [@haspelmath1993: 184]. {#t6 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lk6{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">INCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">EXCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Standard Tabasaran</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uzu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uxu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uču</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">uvu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">učʷu</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Standard Lezgian</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zun</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">čun</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">wun</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">kün</span></td>
  </tr>
</tbody></table>

Two Lezgic languages, Rutul and Kryz, show a certain degree of dialectal variation with respect to the inclusive/exclusive distinction. [Table 7](#t7) shows the pronouns in three varieties of Kryz, namely Kryz Proper (Kryz of Kryz), the Kryz of Alik and Ismailli Kryz. The first two varieties continue the inclusive/exclusive distinction. The Kryz of Alik is notable for its form of the 2PL *jin.* Authier [@authier2009: 43] suggests that the original pronoun (which is *vin* in other dialects) was lost, and the inclusive pronoun *jin* took over its function (see [corresponding section](#dargwa) for similar processes in Dargwa languages).

As for the third variety presented, Ismailli Kryz, it is reported [@authier2021: 216] to have lost the inclusive/exclusive distinction. The originally inclusive pronoun *ji(=n)* took over the functions of the neutral ‘we’.

#### **Table 7.** Personal pronouns of Kryz Proper [@saadiev1994: 420], Alik Kryz [@authier2009: 43] and Ismailli Kryz [@authier2021: 216]. {#t7 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lk6{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">INCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">EXCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Kryz Proper</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zən</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">jin</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žin</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">vən</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">vin</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Alik Kryz</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zi(n)</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">jin</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ži(n)</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">vun</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">jin</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Ismailli Kryz</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zi(=n)</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ji(=n)</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">vun</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">vin</span></td>
  </tr>
</tbody></table>

The situation of the dialects of Rutul is even more complex. [Table 8](#t8) shows the pronouns in some Rutul varieties. Most of them do not distinguish between the inclusive and the exclusive, however, historically different forms are used for 1PL (the same is true for the second person): *je* and *že* / *ži*. In Khnov Rutul the distinction is attested but, similarly to the Archi case, it has been renewed. The inclusive form transparently goes back to the inclusive \**ji* (*jinewi* < *ji=ne wi* [1PL=ADD 2SG], literally ‘we and you’). It is possible that the exclusive *juˁqumbɨ* was also derived from the originally inclusive \**ji* (*juˁqumbɨ* < *juˁ-qu-mbɨ* [1PL-ASSOC-PL], literally ‘we and ours’; cf. in Kina Rutul: *rasul-quner* [Rasul-ASSOC] ‘Rasul and people associated with him’). What is unclear is whether the new inclusive and exclusive pronouns were introduced when \**je* still had the plural meaning, or when it already shifted to the singular: in Khnov, *ji* is used for the 1SG, even though it originally had the meaning of inclusive plural (cf. *je* in Mukhad and Kina Rutul).

Note also the situation in the Mukhad dialect, on which Standard Rutul is based. Mukhad is spoken in the villages of Rutul (the center of the Rutul district), Kufa and Kiche. In Mukhad the inclusive/exclusive distinction is optional. The form *je* can be used both in inclusive and exclusive contexts. However, for the inclusive, speakers tend to use the form *je wɨ*, which is a juxtaposition of the 1PL (*je*) and 2SG (*wɨ*) pronouns. Juxtaposition is not a regular morphological operation in the language, so this construction may be considered a kind of compound reminiscent of the inclusive in Khnov. It then can be concluded that in Mukhad the distinction was also lost, but a new inclusive form (*je wɨ*) is on its way to the pronominal system. As it is not obligatory in inclusive contexts, the villages speaking Mukhad Rutul were assigned the value “not attested”.

#### **Table 8.** Personal pronouns in some Rutul varieties: Mukhad (literary language) (@maxmudova2001: 134; Ilya Sadakov, p.c.), Kina (Ilya Sadakov, p.c.; Timofey Mukhin, p.c.), Shinaz (@ibragimov2004: 133; Ilya Sadakov, p.c.), Myukhrek (@ibragimova2009: 52; Ilya Sadakov, p.c.) and Khnov (@islamov2014: 11; Ilya Sadakov, p.c.). {#t8 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lk6{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">INCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">EXCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Mukhad Rutul</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zɨ</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">je (wɨ)</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">je</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">wɨ</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">waˤ</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Kina Rutul</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zɨ</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">je</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">wɨ</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">we</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Shinaz Rutul</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zɨ</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ži</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʁu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Myukhrek Rutul</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zɨ</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">že / ži</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">wɨ</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Khnov Rutul</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ji</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">jinewi</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">juˁqumbɨ</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʁu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">wi</span></td>
  </tr>
</tbody></table>

One more Rutul dialect that requires a special discussion is the Rutul of Ikhrek; see [Table 9](#t9). According to [@ibragimov2004: 181], there is no inclusive/exclusive distinction. Yet, two forms are used for 1PL — *ži* and *ji* (cf. the form for 1PL *je* in Kina and *ži* in Shinaz). Ibragimov [@ibragimov2004: 181] states that these forms are in free distribution, though *ži* is used more often. Authier [@authier2021: 215] suggests, however, that the Ikhrek dialect actually distinguishes between the inclusive (*ji*) and the exclusive (*ži*). Our field data does not support Authier’s claim and confirms Ibragimov’s statement: both forms *ji* and *ži* were used in both inclusive and exclusive contexts.

#### **Table 9.** Data for Ikhrek Rutul from different sources. {#t9 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lk6{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">INCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">EXCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@ibragimov2004: 181</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zɨ</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ži / ji</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʁu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">Our field data</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zɨ</span></td>
    <td class="tg-7lk6" colspan="2"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ži / ji</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʁu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">@authier2021: 215</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zɨ</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ji</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ži</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʁu</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu</span></td>
  </tr>
</tbody></table>

### Lak and Archi: a case of copy of a rare pattern

The systems of plain personal pronouns in Lak per accounts in [@kibrikkodzasov1990; @zhirkov1955] and in Archi as per [@kibrik1977] are shown in [Table 10](#t10).

#### **Table 10.** Plain personal pronouns in Lak [@kibrikkodzasov1990: 220-222; @zhirkov1955: 65-66] and Archi [@kibrik1977: 125-126]. {#t10 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lk6{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1PL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Literary Lak</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">na</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu</span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent"> (&lt; \*EXCL)</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ina</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zu</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Archi</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zon</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nen</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">un</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žʷen</span></td>
  </tr>
</tbody></table>

In Lak, the first person plural pronoun goes back to the Proto-East-Caucasian exclusive. In Archi, it has several irregular oblique stems, including \**CM-olo* in the genitive and \**CM-el* in the dative (CM stand for a class (gender) marker), and \**la*- followed by case inflections in other cases; the ergative is syncretic with the nominative. The origins of some of these forms, including the nominative-ergative *nen*, are not totally clear [@alekseev1985: 70-73]. 
What is more important for the purposes of this survey, in Archi all these forms have strictly exclusive reference. To convey the meaning of the inclusive, they must combine with the particle -*a-CM-u* (CM stand for a class (gender) marker; the particle itself has a wide range of functions from deriving reflexive pronouns to the obligatory use with cardinal numerals) [@mikailov1967: 84; @kibrik1977: 125-126]. For standard Lak, [@friedman2021] notes that there is the same tendency to use plain first person pronouns for the exclusive and to add what he calls emphatic/reflexive particle for the inclusive reference. Our personal data suggests that it is a strict distribution in Lak of Kuba, though further study and data from other varieties is required. Cf. [Table 11](#t11) (where -*t'-u* should be interpreted as < \**-a<t’>u*):

#### **Table 11.** Exclusive and inclusive pronouns in Archi and Lak. {#t11 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lvb{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-reig{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:left;vertical-align:top}
.tg .tg-tyfs{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:left;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev" rowspan="2"></th>
    <th class="tg-hwu2" colspan="2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Archi</span></th>
    <th class="tg-hwu2" colspan="2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Lak</span></th>
  </tr>
  <tr>
    <th class="tg-7lvb"><span style="font-style:normal">INCL</span></th>
    <th class="tg-7lvb"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">EXCL</span></th>
    <th class="tg-7lvb"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">INCL</span></th>
    <th class="tg-7lvb"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">EXCL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-reig"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Nom</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nen</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">we</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">nen-t’u</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">we-ɪᴠ.ᴘᴛᴄʟ</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">we</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu-wa</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">we-ɪᴠ.ᴘᴛᴄʟ</span></td>
  </tr>
  <tr>
    <td class="tg-reig"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Gen</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">olu</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">ɪᴠ.we.ɢᴇɴ</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">la-t’u</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">ɪᴠ.we.ɢᴇɴ-ɪᴠ.ᴘᴛᴄʟ</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu-l</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">we-ɢᴇɴ</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu-l-a</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">we-ɢᴇɴ-ɪᴠ.ᴘᴛᴄʟ</span></td>
  </tr>
  <tr>
    <td class="tg-reig"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Dat</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">el</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">ɪᴠ.we.ᴅᴀᴛ</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">el-t’u</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">ɪᴠ.we.ᴅᴀᴛ-ɪᴠ.ᴘᴛᴄʟ</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu-n</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">we-ᴅᴀᴛ</span></td>
    <td class="tg-tyfs"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">žu-n-a</span><br><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent">we-ᴅᴀᴛ-ɪᴠ.ᴘᴛᴄʟ</span></td>
  </tr>
</tbody></table>

The parallel between the Lak and Archi pronouns is all the more notable as we are unaware of cases of morphological derivation of the inclusive pronoun from the unmarked exclusive pronoun in other languages of the world. Both Lak and Archi particles carry gender markers, but their cognacy is questionable. Archi, a Lezgic outlier in close contact with Lak [@chumakina2009; @dobrushina2013; @daniel2026] must have copied the pattern from its influential neighbour, as was first suggested in [@kaxadze1964], see also [@antonenko2011]. 

We conclude that Lak and Archi have lost their original inclusive/exclusive distinction, but Lak has subsequently renewed the distinction recruiting a new morphological means; and Archi followed in its wake.

### The other branches: Nakh, Tsezic and Khinalug

Tsezic languages lack the inclusive/exclusive distinction.[Table 12](#t12) shows the personal pronouns of Bezhta (Tsezic) as spoken in the village of Khasharkhota. The origin of the neutral ‘we’ in Tsezic languages is the proto-East-Caucasian inclusive form \**Lǟ* [@nikolaevstarostin1994].

#### **Table 12.** Personal pronouns in Bezhta [@kibrikkodzasov1990: 220-222]. {#t12 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lvb{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;font-weight:bold;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1PL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Bezhta, Khasharkhota village</span></td>
    <td class="tg-7lvb"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">do</span></td>
    <td class="tg-7lvb"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ile</span><span style="font-weight:400;font-style:normal;text-decoration:none;background-color:transparent"> (&lt; *1INCL)</span></td>
    <td class="tg-7lvb"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">mi</span></td>
    <td class="tg-7lvb"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">miže</span></td>
  </tr>
</tbody></table>

On the other hand, the distinction is attested in Nakh and Khinalug. The personal pronouns of Khinalug and Chechen (Nakh) are given in [Table 13](#t13).

#### **Table 13.** Khinalug [@kibrikkodzasov1990: 220-223 ; @kibriketal1972: 72] and Chechen personal pronouns [@nichols1994: 32]. {#t13 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;margin:0px auto;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-hwu2{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-p5ev{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  text-align:left;vertical-align:top}
.tg .tg-7lvb{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;font-weight:bold;text-align:center;vertical-align:top}
.tg .tg-7lk6{border-color:#ffffff;color:#555555;font-family:"Charis SIL", sans-serif !important;font-size:14px;
  font-style:italic;text-align:center;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-p5ev"></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1INCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">1EXCL</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2SG</span></th>
    <th class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">2PL</span></th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Khinalug</span></td>
    <td class="tg-7lvb"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zɨ</span></td>
    <td class="tg-7lvb"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">kin</span></td>
    <td class="tg-7lvb"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">jir</span></td>
    <td class="tg-7lvb"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">vɨ</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">zur</span></td>
  </tr>
  <tr>
    <td class="tg-hwu2"><span style="font-weight:700;font-style:normal;text-decoration:none;background-color:transparent">Chechen</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">swo</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">vaj</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">txwo</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">ʜwo</span></td>
    <td class="tg-7lk6"><span style="font-weight:400;font-style:italic;text-decoration:none;background-color:transparent">šu</span></td>
  </tr>
</tbody></table>

## Distribution

The following general pattern can be seen on the [map]. The languages that distinguish between the inclusive and the exclusive are located in the west, whereas the ones that do not make the distinction are situated in the east of Daghestan. There are two main exceptions to this pattern: Tsezic languages in the west, which do not have the distinction, and the area in the center, where Tabasaran, Agul and some Dargwa varieties are spoken, which lack the distinction.

This general areal pattern can be by and large explained by the genealogical distribution of the languages: the ones which have the distinction, i.e. Avar-Andic, are spoken in the northwest of Dagestan, while those that mostly lack it, i.e. Lezgic languages, are spoken in the southeast. Dargwa languages are geographically split into the southern area, where the distinction is present at least in some varieties, and the northern area, where it is completely absent. The southern Dargwa language area is adjacent to Agul and Tabasaran, both Lezgic, which preserved the distinction.

As suggested in [@authier2021], the languages in the geographical center (Lak, some Dargwa in the South, Agul and Tabasaran) preserve the distinction between inclusive and exclusive inherited from the proto-languages, whereas the languages in the periphery, which are more prone to contacts with non-East-Caucasian languages, like Azerbaijani, Georgian, Kumyk, etc., lack it. The loss of the inclusive in Tsezic is explained by Authier by relatively weak contact with Andic languages and strong contact with Georgian. However, Tsezic is in contact with Avar, which has the opposition [@dobrushinaetal2017].

To conclude, the distribution of the inclusive/exclusive distinction is primarily explained by genealogy. However, some influence of geographical factors may also be suspected, including contacts with non-East-Caucasian languages which lack the distinction. This hypothesis requires further investigation.

```{r, include=FALSE}
gloss_example("a b c d",
              "PTCL GEN DAT IV OBL",
              "FAKE CALL")
```


## List of glosses {-}

```{r}
make_gloss_list(readxl::read_xlsx("../glosses.xlsx"))
```

## References {-}
