```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = '')
library(lingglosses)
```

## Introduction

In this chapter, the term *jussive* is used for forms or constructions which express commands directed towards someone who is not a speech act participant, such as English *Let them all come!* The addressee serves as a mediator between the speaker and this participant. This category is sometimes called *3rd person imperative* [@xrakovskij2001; @aikhenvald2010] or *exhortative* [@ammannvanderauwera2004].

The central meaning of jussive forms is indirect command:

(@one) Kryz [@authier2009: 281]
```{r}
gloss_example(transliteration = "va i-nkan-i pul **tu-ğa-tir**!",
              glosses = "you.SG.GEN PV-IPFV.remain-PTCP money **PV-carry-JUSS**",
              free_translation = "Let him bring back the rest of your money.")
```

Additionally, jussives usually can express permission (@two) and indifference (@three):

(@two) Archi [@dobrushina2012: 11]
```{r, results='asis'}
gloss_example(transliteration = "sa‹r›ka laha guruč’an orχir-ši —  **χːeχːa-ba**",
              glosses = "‹II›look(IMP) child.OBL(ERG) sweet(IV).NOM IV.take.IPFV-CVB — **IV.take.IMP-JUSS**",
              free_translation = "Look, the child is taking a sweet. — Let him take it.")
```

(@three) Mehweb Dargwa [@dobrushina2019: 136]
```{r}
gloss_example(transliteration = "**rurž-e** **bet'-a** ħarši",
              glosses = "**boil.IPFV-IMP** **say.PFV-IMP.TR** soup",
              free_translation = "Let the soup boil. [I don't care / Or: It needs more time to simmer]")
```

Finally, in some languages jussives are also used to express wishes, in particular blessings and curses (@five) - a type of speech act which is very common in everyday communication in the Caucasus:

(@four) Kumyk - third person command (N. E. Gadzhiaxmedov, p.c., cited in [@dobrushina2012])
```{r}
gloss_example(transliteration = "mahammat üj-ge bar-ɨp dars-lar-ɨ-n **et-sin**",
              glosses = "Mohammad house-DAT go-CVB lesson-PL-POSS.3-ACC **make-JUSS**",
              free_translation = "Let Mohammad go home and do his homework.")
```

(@five) Kumyk - third person blessings and curses (N. E. Gadzhiaxmedov, p.c., cited in [@dobrushina2012])
```{r}
gloss_example(transliteration = "ömür-lü **bol-sun**",
              glosses = "long_life-ADJ **be-JUSS**",
              free_translation = "Let him be a long-living (person). [A blessing uttered to parents of a newborn baby, addressed to the latter]")
```

Some Nakh-Daghestanian languages have separate forms for indirect commands and for wishes (for example, Mehweb, Archi, Lak, and Chirag Dargwa), but the co-expression of these two meanings by the same form is more frequent. Hence it might be difficult to distinguish jussive from *optative*, a form whose main function is to express a wish or hope of the speaker that something would happen [@dobrushina2021]. Many Nakh-Daghestanian languages, such as Agul and Rutul, have the same form to express blessings and curses addressed to 1st, 2nd or 3rd persons, and to express 3rd person imperative. This form cannot be used to urge 1st or 2nd persons to do something (see examples in [@dobrushina2021]). In this chapter, forms of the Agul and Rutul type will be classified as *optatives*. I reserve the term *jussive* for forms that are restricted to 3rd person imperative (as in Mehweb, see [@dobrushina2019: 133–138], or combine 3rd person imperative with the expression of a wish with a 3rd person subject (as in Kumyk, see [@gadzhiakhmedov2000]). The terms are presented in [Table 1](#t1) (see also the chapter on [Optatives](http://lingconlab.ru/dagatlas/007_optative.html), and the discussion of these terms in [@dobrushina2012]).

##### **Table 1.** Optative and jussive (from [@dobrushina2021]) {#t1 -}

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-llyw{background-color:#c0c0c0;border-color:inherit;text-align:left;vertical-align:top}
.tg .tg-0pky{border-color:inherit;text-align:left;vertical-align:top}
</style>
<table class="tg">
<thead>
  <tr>
    <th class="tg-0pky"></th>
    <th class="tg-0pky" colspan="2">Optative<br>(Mehweb form in -b)</th>
    <th class="tg-0pky" colspan="2">Optative<br>(Rutul form in -Vj)</th>
    <th class="tg-0pky" colspan="2"><span style="font-weight:400;font-style:normal;text-decoration:none">Jussive</span><br><span style="font-weight:400;font-style:normal;text-decoration:none">(Kumyk form in –syn/-sin/-sun)</span></th>
    <th class="tg-0pky" colspan="2">Jussive<br>(Mehweb form with bet’a)</th>
  </tr>
</thead>
<tbody>
  <tr>
    <td class="tg-0pky">person</td>
    <td class="tg-0pky">command</td>
    <td class="tg-0pky">blessings and <br>curses</td>
    <td class="tg-0pky">command</td>
    <td class="tg-0pky">blessings and <br>curses</td>
    <td class="tg-0pky">command</td>
    <td class="tg-0pky">blessings and <br>curses</td>
    <td class="tg-0pky">command</td>
    <td class="tg-0pky">blessings and <br>curses</td>
  </tr>
  <tr>
    <td class="tg-0pky">1</td>
    <td class="tg-0pky"></td>
    <td class="tg-llyw"></td>
    <td class="tg-0pky"></td>
    <td class="tg-llyw"></td>
    <td class="tg-0pky"></td>
    <td class="tg-0pky"></td>
    <td class="tg-0pky"></td>
    <td class="tg-0pky"></td>
  </tr>
  <tr>
    <td class="tg-0pky">2</td>
    <td class="tg-0pky"></td>
    <td class="tg-llyw"></td>
    <td class="tg-0pky"></td>
    <td class="tg-llyw"></td>
    <td class="tg-0pky"></td>
    <td class="tg-0pky"></td>
    <td class="tg-0pky"></td>
    <td class="tg-0pky"></td>
  </tr>
  <tr>
    <td class="tg-0pky">3</td>
    <td class="tg-0pky"></td>
    <td class="tg-llyw"></td>
    <td class="tg-llyw"></td>
    <td class="tg-llyw"></td>
    <td class="tg-llyw"></td>
    <td class="tg-llyw"></td>
    <td class="tg-llyw"></td>
    <td class="tg-0pky"></td>
  </tr>
</tbody>
</table>

<br>

## Results

This chapter takes into account all formal types of jussives: those which contain a dedicated affix, those which are expressed by some other grammatical category (usually optative), and those which are expressed by a periphrastic construction: combination of some form with another verb. The maps answer the following questions:

- Is there a dedicated jussive marker in the given language ([Map 1](http://lingconlab.ru/dagatlas/055_jussive_map.html#m1))?
- Is there a form or construction which expresses jussive meaning and has a verb 'say' as its source ([Map 2](http://lingconlab.ru/dagatlas/055_jussive_map.html#m2))?
- Is there a jussive encoded by morphological causative markers ([Map 3](http://lingconlab.ru/dagatlas/055_jussive_map.html#m3))?
- Is there an optative form which co-expresses jussive meaning ([Map 4](http://lingconlab.ru/dagatlas/055_jussive_map.html#m4))?

Note that a language can get positive answers for more than one of these questions. For example,  Kryz has two forms expressing jussive meaning -  proper jussive, restricted to 3rd person (*votif* [@authier2009: 281]) and optative with jussive meaning (*optatif* [@authier2009: 279–280]).

[Map 1](http://lingconlab.ru/dagatlas/055_jussive_map.html#m1) reflects the presence of a dedicated affix for jussive in the languages of Daghestan. Besides Turkic languages (see (@four) and (@five)), dedicated morphological jussives are quite infrequent in the area, and often seem innovations with recently grammaticalized suffixes. The Itsari jussive with the suffix *-ikːa* is supposed to derive from the verb 'want' [@mutalov2002: 131] - a unique pattern for the area. Lak, to the contrary, features a pattern that is frequently found in Daghestan - it encodes jussive with the suffix derived from the imperative of the verb 'say':

(@six) Lak [@eldarova1999: 57]
```{r}
gloss_example(transliteration = "uča lasi-ča",
              glosses = "say.IMP take.IMP-say.IMP",
              free_translation = "Let them take.")
```

Since the imperative of the Lak verb 'say' underwent reduction and is not identical to the jussive suffix, Lak jussive is qualified as morphologically dedicated. Only Tsakhur and Kryz (@one) have jussive suffixes whose origin is not clear.

Grammaticalization of the verb 'say' to the jussive marker is attested in many languages, although not everywhere it has reached the stage of a suffix as in Lak. For example, Mehweb jussive is expressed by a combination of the full imperative of the main verb and the imperative of the verb 'say'. The fact that the main arguments in the jussive construction are marked in accordance with the lexical verb, and not with the verb 'say', shows that this is not a regular combination of the matrix verb 'say' and an embedded imperative. Compare the following examples. In (@seven a) *Musa* is in the nominative, since the main argument of the verb 'work' is normally in the nominative. In (@seven b) Muħammad stands in the ergative, because 'sing' is a transitive verb with ergative-nominative alignment. Finally, in (@seven c) we observe a usual (non-jussive) construction with the verb 'say', and the addressee is marked by inter-lative.

(@seven) Mehweb [@dobrushina2019: 135]
a.
```{r}
gloss_example(transliteration = "musa **uz-e** **bet'-a**",
              glosses = "Musa **M.work.IPFV-IMP** **say.PFV-IMP.TR**",
              free_translation = "Let Musa work.")
```
b.
```{r}
gloss_example(transliteration = "muħammad-ini deč' **b-aq'-a** **bet'-a**",
              glosses = "Muhammad-ERG song **N-do.PFV-IMP.TR** **say.PFV-IMP.TR**",
              free_translation = "Let Muhammad sing a song.")
```
c.
```{r}
gloss_example(transliteration = "musa-ze **uz-e** **bet'-a**",
              glosses = "Musa-INTER(LAT) **M.work.IPFV-IMP** **say.PFV-IMP.TR**",
              free_translation = "Tell Musa to work.")
```

[Map 2](http://lingconlab.ru/dagatlas/055_jussive_map.html#m2) shows all 'say'-based jussives, regardless of their degree of grammaticalization - from affixes (as in Lak) to the periphrastic construction with two verbs, as in Mehweb, Archi, and Chirag Dargwa. Note that this map covers not only jussives, but also optatives with jussive functions as, for example, in Hinuq or Chamalal. In the chapter on [Optatives](http://lingconlab.ru/dagatlas/007_optative.html) [@dobrushina2021] there is another map which covers only 'say'-optatives.

Much less frequently is jussive expressed as the imperative of a causativized verb. This pattern is found in Chechen, Ingush, and Kubachi Dargwa.

(@eight) Chechen (Zarina Molochieva, p.c., cited in [@dobrushina2012])
```{r}
gloss_example(transliteration = "p'eːt'mat-ie beːpig d-a-it-a",
              glosses = "Patimat-ALL bread(III).NOM III-make-CAUS-IMP",
              free_translation = "Let Patimat bake bread / Make Patimat bake bread.")
```

[Map 3](http://lingconlab.ru/dagatlas/055_jussive_map.html#m3) reflects only jussives encoded by morphological causative markers. Note that there are also many languages which use some kind of periphrastic constructions with the causative verb 'let' or a particle derived from this verb, but these cases were not included in the map, because it is usually difficult to distinguish between obligatory marking by causative verbs or particles and optional marking, which is found in many languages.

[Map 4](http://lingconlab.ru/dagatlas/055_jussive_map.html#m4) compares proper jussives and optatives which co-express jussive functions. As was mentioned in the introduction, optatives, unlike jussives, are compatible with all three persons. In some languages, such as Agul, Bagvalal, Tanty Dargwa, and Karata, optatives express an indirect command when they are used with a third person subject; these languages are reflected on this map as "optative". Some other languages, such as Mehweb, have separate forms for jussive and optative, the first being dedicated to the expression of an indirect command, the second being dedicated to the expression of blessings and curses. Such languages are shown as "non-optatives".

Note, however, that in many languages the distinction between the form with jussive function (indirect command) and the form with optative functions (wishes) can be observed only on the level of frequency. For example, Kryz has two forms that can be used both for indirect command and for wishes, but the first function is much more typical for the jussive (*votif*), while the second one is more typical for the optative, according to Authier [-@authier2009]. In the absence of large corpora such issues cannot be solved, so in this study we classify the forms based mainly on the researchers' judgments.

## Distribution

A dedicated jussive affix is found in all Turkic languages (Kumyk, Nogai, and Azerbaijani), Itsari, Lak and two Lezgic languages, Kryz and Tsakhur. Since both Kryz and Tsakhur are in contact with Azerbaijani, one can suppose that the dedicated jussive has developed under its influence (as was suggested in [@authier2010]).

'Say'-affixes and constructions are especially typical of Tsezic and Andic languages, but are also found in some Dargwa varieties and in Lak. Among Lezgic languages, only Rutul features this form; its geographical position and known history of language contact does not allow to hypothesize the contact origin of this form. There is also one Indo-European language which has an optative/jussive form with a 'say'-suffix, namely Tat.

Morphological causatives encode jussive meanings very infrequently. This is found in two Nakh languages (Ingush and Chechen) and, unexpectedly, in Kubachi Dargwa.

Co-expression of jussives by optatives is by far the most frequent way to express jussive meanings in the Nakh-Daghestanian family. There are, however, some notable exceptions. These are Ingush, Chechen, Avar, Botlikh, Lak, Kubachi Dargwa, Mehweb Dargwa, Chirag Dragwa, Itsari Dargwa, and Tsakhur; some of these languages have several optative affixes, none of which is used for jussives.

## List of glosses {-}

```{r}
make_gloss_list(readxl::read_xlsx("../glosses.xlsx"))
```

## References {-}
