

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning=FALSE, fig.width = 9.5)
library(tidyverse)
library(lingtypology)
library(RefManageR)
```

## {-}

Visualization 1 (**Extrapolated data**) and 3 (**General datapoints**) are both based on the principle 'one value – one language'. Visualization 2 (**Data granularity**) takes into account dialect levels. On the **Data granularity** maps you can see the village and the data type (village data, general language data, etc.) when you click on a dot. Hover over or click on a dot to see the language. By unticking the box “show languages” you can remove the inner dots and visualize the distribution of different values in the area without the distraction of genealogical information.

```{r}

feature_dataset <- read_tsv("../orig_table/laterals.tsv")


feature_dataset %>% 
  filter(map == 'yes') ->
  feature_dataset_4map

villages <- read_tsv('../tald_villages.csv') # village coordinates
genlang <- read_tsv('../genlangpoints.csv') # general language points

feature_dataset_4map %>% 
  filter(type == 'language') ->
  feature_dataset_4map_languages

feature_dataset_4map %>% 
  filter(type != 'language') ->
  feature_dataset_4map_rest

genlang %>% 
  filter(aff == 'Dargwa',
         lang != 'Kaitag',
         lang != 'Standard Dargwa') ->
  change

if(nrow(change) > 0) {
  villages$gltc_lang[which(villages$village_dialect %in% change$gltc_lang
)] <- change$gltc_lang

  villages$gltc_lang[which(villages$dialect_toplevel == 'Kaitag')] <- 'kajt1238'
}

villages %>% 
  select(village, rus_village, lat, lon, gltc_lang, lang, aff, family, standard, dialect_toplevel, dialect_nt1, dialect_nt2, dialect_nt3, village_dialect, lang_col, aff_col) %>% 
  pivot_longer(names_to = 'type', values_to = 'idiom', standard:village_dialect) %>% 
  filter(!is.na(idiom)) %>% 
  mutate(type = case_when(type == 'village_dialect' ~ 'village',
                          TRUE ~ type)) %>%
  inner_join(feature_dataset_4map_rest, by = c('type', 'idiom', 'lang')) %>% 
  distinct() ->
  all_data_without_languages

villages %>% 
  select(village, rus_village, lat, lon, gltc_lang, lang, aff, family, lang_col, aff_col) %>%
  filter(lang %in% feature_dataset_4map_languages$lang) %>%
  anti_join(all_data_without_languages %>%  select(village)) %>% 
  inner_join(feature_dataset_4map_languages) %>% 
  distinct() %>% 
  bind_rows(all_data_without_languages)  ->
  alldata_clean  

alldata_clean  %>% 
  distinct(gltc_lang) %>% 
  mutate(lang4map = lang.gltc(gltc_lang),
         display = 'show languages') %>% 
  right_join(alldata_clean) %>% 
  filter(!is.na(contributor)) %>% 
  mutate(type = factor(type, levels = c('language', 'dialect_toplevel', 'dialect_nt1', 'dialect_nt2', 'dialect_nt3', 'village'))) ->
  alldata_clean

#make short

villages %>% 
  filter(is.na(id)) %>%
  select(village, rus_village, lat, lon, gltc_lang, lang, aff, family, lang_col, aff_col) %>%
  left_join(feature_dataset_4map_languages) %>% 
  distinct() %>% 
  bind_rows(all_data_without_languages) %>% 
  mutate(lang4map = lang.gltc(gltc_lang),
         display = 'show languages') %>% 
  filter(!is.na(contributor))  ->
  all_genpoints
  
rm(all_data_without_languages, change, feature_dataset_4map, feature_dataset_4map_languages, feature_dataset_4map_rest, villages, genlang)
```

## Presence of lateral obstruents{.tabset .tabset-fade .tabset-pills #m1} 
    
    
### Extrapolated data {-}

```{r}
map.feature(alldata_clean$lang4map,
            latitude = alldata_clean$lat, 
            longitude = alldata_clean$lon,
            features = as.factor(alldata_clean$value1),
            color = 'magma',
            title = alldata_clean$value1_name[1],
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = alldata_clean$lang,
            zoom.control = TRUE,
            width = 8,
            popup = paste(alldata_clean$village, '|',
                          alldata_clean$rus_village, '<br>',
                          'data:', alldata_clean$type)) %>% 
    map.feature(alldata_clean$lang4map,
              latitude = alldata_clean$lat,
              longitude = alldata_clean$lon,
              features = alldata_clean$lang,
              color = alldata_clean$lang_col,
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = alldata_clean$display,
              popup = paste(alldata_clean$village, '|',
                        alldata_clean$rus_village, '<br>',
                          'data:', alldata_clean$type))
```

### Data granularity {-}

```{r}
map.feature(alldata_clean$lang4map,
            latitude = alldata_clean$lat, 
            longitude = alldata_clean$lon,
            features = as.factor(alldata_clean$value1),
            color = 'magma',
            title = alldata_clean$value1_name[1],
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = alldata_clean$lang,
            zoom.control = TRUE,
            control = alldata_clean$type,
            popup = paste(alldata_clean$village, '|',
                        alldata_clean$rus_village, '<br>',
                          'data:', alldata_clean$type))
```

### General datapoints {-}

```{r}
map.feature(all_genpoints$lang4map,
            latitude = all_genpoints$lat,
            longitude = all_genpoints$lon,
            features = factor(all_genpoints$value1),
            title = all_genpoints$value1_name[1],
            color = 'magma',
            legend = TRUE,
            legend.position = 'bottomleft', 
            zoom.control = TRUE,
            width = 8) %>% 
  map.feature(all_genpoints$lang4map,
              latitude = all_genpoints$lat,
              longitude = all_genpoints$lon,
              features = all_genpoints$lang,
              color = all_genpoints$lang_col,
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = all_genpoints$display)
```
    


## Datatable

```{r}
bib <- RefManageR::ReadBib(file = '../bibliography.bib')

tibble(column = str_subset(colnames(feature_dataset), 'value\\d{1,}$')) %>%
  mutate(new_name = feature_dataset %>% 
           select(str_c(column, '_name')) %>% 
           unlist() %>% 
           unique()) ->
  columns_rename

feature_dataset %>% 
  select(lang, idiom, source, page, matches('value\\d{1,}$')) %>% 
  rename_with(function(x){columns_rename[columns_rename$column == x, ]$new_name}, matches('value\\d{1,}$'))  %>% 
  rename(Language=lang, 
         Idiom = idiom,
         Source = source) %>% 
  mutate(page = str_replace_all(page, '--', '–'),
         Source = str_split(Source, '; '),
         page = str_split(page, '; ')) %>% 
  unnest_longer(col = c(Source, page)) %>% 
  rowwise() %>% 
  mutate(page = ifelse(page == 'NA', NA_character_, page),
         Source = ifelse(str_detect(Source, 'p.c.$'),
                         Source,
                         Cite(bib, Source, 
                       after = ifelse(!is.na(page), 
                                      str_c(': ', page),
                                      '')))) %>%
  select(-page) %>% 
  DT::datatable(class = 'cell-border stripe', 
    rownames = FALSE, 
    filter = 'top', 
    extensions = 'Buttons',
    options = list(pageLength = 100, 
                   autoWidth = TRUE, 
                   info = FALSE,
                   dom = 'fBltp',
                   buttons = list(list(extend = 'collection',
                                       buttons = c('csv', 'excel', 'pdf'),
                                       text = '<i class="fas fa-download"></i>')),
                   paginate = TRUE))
```

## References {-}


