---
output: html_document
#bibliography: numeral_morphology.bib
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(lingglosses)
library(tidyverse)
library(bib2df)

# you need to provide the real name, not `chapter_template`
#file_name <- "numeral_morphology_bib.tsv"
#read_tsv(file_name, progress = FALSE) %>% 
#  df2bib(file_name %>% str_remove_all("[_\\.]tsv") %>%  str_replace("_bib$", "\\.bib"))
```

# Presence of the numeral marker in ordinals

## 1. Introduction

This chapter is dedicated to presence or absence of the numeral marker in ordinal numerals across the East Caucasian family.

## 2. Results
Ordinals express the index of an object in a row of other objects, e.g. English ‘fourth’. Ordinals have their own suffixes, but in some East Caucasian languages these suffixes also contain numeral markers:

(@) Karata proper [@magomedbekova1971: 96]
```{r, echo=FALSE}
gloss_example(transliteration = "ɬab-**da-ƛo-b**",
              glosses = "three-**NUM-ORD-CM**",
              free_translation = "third"
)
```

In others, the numeral marker is absent:

(@) Tindi [@magomedova2012: 145]
```{r, echo=FALSE}
gloss_example(transliteration = "ɬabi-liƛa-b kalasi",
              glosses = "three-ORD-CM class",
              free_translation = "third class"
)
```

In East Caucasian languages, ordinal markers commonly originate in past participles from the verb ‘say’ with different aspectual features, as shown by [@nasledskovanetkachev2021; @martzloff2017: 156]. This feature is not mapped in WALS, though it includes  a map on expression of the first three natural numerals in the ordinal series [@stolzveselinova2013].

### 3. Areal distribution
In Dargic languages ordinals usually do not have a numeral marker, while in Lezgic and Tsezic languages they usually do. Apart from that, this feature is rather inconsistent across branches.

## List of glosses
`r add_gloss("2")`— second person; `r add_gloss("ABS")` — absolutive; `r add_gloss("ADD")`, `r add_gloss("ADD1")`, `r add_gloss("ADD2")` — additive; `r add_gloss("ADJ")` — adjective; `r add_gloss("ATTR")` — attributive; `r add_gloss("CM")` — class marker; `r add_gloss("COM")` — comitative; `r add_gloss("D")` — class D; `r add_gloss("DEM")` — demonstrative; `r add_gloss("DISTR")` — distributive; `r add_gloss("EMPH")` — emphatic particle; `r add_gloss("ERG")` — ergative; `r add_gloss("I")` — class I; `r add_gloss("III")` — class III; `r add_gloss("INT")` — intensifier; `r add_gloss("IV")` — class IV; `r add_gloss("J")` — class IV; `r add_gloss("LAT")` — lative; `r add_gloss("LNK")` — linking suffix; `r add_gloss("M")` — masculine class; `r add_gloss("N")` — neuter class; `r add_gloss("N2")` — second neuter class; `r add_gloss("NUM")` — numeral marker; `r add_gloss("OBL")` — oblique; `r add_gloss("ORD")` — ordinal; `r add_gloss("PFV")` — perfective; `r add_gloss("SG")` — singular

## References
