
## Introduction

This chapter is dedicated to presence or absence (or inclusion) of a class suffix in numeral markers across the East Caucasian family.

## Results
Class suffixes play the role of numeral markers in Lezgic languages. Unlike intensifying suffixes in Avar-Andic languages and -*al* in Dargic languages, class suffixes are generally not absent in complex numerals. However, usually only the marker on the final component agrees with the corresponding noun. Other markers are frozen and have the default value for abstract objects (usually class IV or non-human):

(@budukh_35_rubles) Budukh [@talibov2007: 152]
```{r, echo=FALSE}
gloss_example(transliteration = "q’a-n-na ɨs-na fu-b manat",
              glosses = "twenty-IV-ADD ten-ADD five-III ruble.III",
              free_translation = "thirty five rubles"
)
```

It is debatable whether these class markers should be interpreted as numeral markers. One plausible argument against this could be that class marking is a typical feature for adnominal elements in Lezgic languages, and the markers here resemble class markers in adjectives. Therefore, they are not numeral markers, but rather adnominal markers, indicating agreement with the noun in different word classes. However, at least for some Lezgic languages the behavior of these markers in inflection differs in adjectives and numerals. For instance, in Rutul adjectives generally do not inflect in class and do not agree with the noun in case, whereas numerals agree with the noun in absolutive noun phrases. Consider the following examples:

(@) Mukhad Rutul [@ibragimov2004: 73--76]
```{r, echo=FALSE}
gloss_example(transliteration = "q’ux-du did",
              glosses = "big-ADJ father",
              free_translation = "big father (absolutive)"
)
```
```{r, echo=FALSE}
gloss_example(transliteration = "q’ux-du did-æ",
              glosses = "big-ADJ father-ERG",
              free_translation = "big father (ergative)"
)
```
```{r, echo=FALSE}
gloss_example(transliteration = "xib-**ɨr** gada",
              glosses = "three-**I** boy(I)",
              free_translation = "three boys (absolutive)"
)
```
```{r, echo=FALSE}
gloss_example(transliteration = "xib-**dɨ** gadij-eræ",
              glosses = "three-**OBL** boy(I)-ERG",
              free_translation = "three boys (ergative)"
)
```

Some grammars (for example, [@nichols2011: 197], [@mukhin2020: 6], [@kibrik1999: 157]) claim that class markers agree with higher numerals like ‘hundred’ or ‘thousand’ and that is why class markers before these numerals are always frozen in a specific class. However, this does not explain the presence of a class marker in the numeral ‘20’ in the example from Budukh (@budukh_35_rubles).

In Lezgian, the nominal class category is lost, hence, class markers in numerals are petrified as suffixes. In Agul, the petrified class marker appears in headless contexts only, hence I treat Agul as a language which does not have a numeral marker.

In Lak and Archi, numeral markers also contain a class marker. However, the numeral markers are absent in non-final positions in complex numerals so there are no frozen class markers in these languages. The same is true for Akhvakh, as shown in (@n_akhvakh_40). Akhvakh is the only Andic language where a class suffix as part of the numeral marker is attested. However, there is some confusion in the sources regarding the distribution of this feature across dialects.

(@n_akhvakh_40) Northern Akhvakh [@magomedbekova1967: 40]
```{r, echo=FALSE}
gloss_example(transliteration = "k’e-q’endo-**da-be**",
              glosses = "two-twenty-**NUM-CM**",
              free_translation = "40"
)
```

In some languages, class slots are integrated into some of the numeral roots, where they appear in a non-final position. For example, this is the case with ‘4’ in Ingush (note the uppercase letters which denote class):

(@) Standard Ingush [@nichols2011: 197--198]
```{r, echo=FALSE}
gloss_example(transliteration = "**j**.iʔ jiʃa",
              glosses = "**J**.four {sister(J)}",
              free_translation = "four sisters"
)
```
```{r, echo=FALSE}
gloss_example(transliteration = "**d**.iʔ sahwat",
              glosses = "**D**.four {hour(D)}",
              free_translation = "four hours"
)
```

Such class slots are not taken into consideration in this study. They appear in a small subset of East Caucasian languages, and in those languages there are not more than a few roots which have such slots. These markers are clearly inseparable from the root and do not appear after the root. They are used together with the final class suffixes without contradiction. Thus, they appear to be a separate phenomenon.

## Distribution
Class slots appear as a part of the numeral marker in Lak and adjacent Archi, as well as in Akhvakh dialects. Numeral markers are class suffixes in Lezgic languages. In other branches, there are no class slots in numeral markers.

## List of glosses
`r add_gloss("ADD")` — additive; `r add_gloss("ADJ")` — adjective; `r add_gloss("CM")` — class marker; `r add_gloss("D")` — class D; `r add_gloss("ERG")` — ergative; `r add_gloss("I")` — class I; `r add_gloss("III")` — class III; `r add_gloss("IV")` — class IV; `r add_gloss("J")` — class J; `r add_gloss("NUM")` — numeral marker; `r add_gloss("OBL")` — oblique

## References
