---
output: html_document
#bibliography: numeral_morphology.bib
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(lingglosses)
library(tidyverse)
library(bib2df)

# you need to provide the real name, not `chapter_template`
#file_name <- "numeral_morphology_bib.tsv"
#read_tsv(file_name, progress = FALSE) %>% 
#  df2bib(file_name %>% str_remove_all("[_\\.]tsv") %>%  str_replace("_bib$", "\\.bib"))
```

# Presence of the numeral marker in the numeral ‘1 000’

## 1. Introduction

This chapter is dedicated to the presence of the numeral marker in the numeral ‘1 000’ and other thousands across the East Caucasian family and neighboring languages.

## 2. Results
The numerals with the root ‘1 000’ can vary in regard to the presence of the marker as well as hundreds. In Bagvalal, similarly to crowns and hundreds, only the numeral ‘1 000’ bears the marker whereas numerals ‘2 000’—‘9 000’ do not:

(@) Bagvalal [@tatevosov2001numerals: 155]:
```{r, echo=FALSE}
gloss_example(transliteration = "azar-da",
              glosses = "thousand-NUM",
              free_translation = "1 000"
)
```
```{r, echo=FALSE}
gloss_example(transliteration = "hač’ʷa:-ʔazar",
              glosses = "nine-thousand",
              free_translation = "9 000"
)
```


### 3. Areal distribution
Patterns of usage of the marker generally correspond to the large branches and are very similar to those for ‘100’. In Lezgic, Dargic, and Tsezic languages, the marker is usually absent, in Avaro-Andic languages it is present. It is also present in Lak and adjacent Archi of the Lezgic branch.

## List of glosses
`r add_gloss("2")`— second person; `r add_gloss("ABS")` — absolutive; `r add_gloss("ADD")`, `r add_gloss("ADD1")`, `r add_gloss("ADD2")` — additive; `r add_gloss("ADJ")` — adjective; `r add_gloss("ATTR")` — attributive; `r add_gloss("CM")` — class marker; `r add_gloss("COM")` — comitative; `r add_gloss("D")` — class D; `r add_gloss("DEM")` — demonstrative; `r add_gloss("DISTR")` — distributive; `r add_gloss("EMPH")` — emphatic particle; `r add_gloss("ERG")` — ergative; `r add_gloss("I")` — class I; `r add_gloss("III")` — class III; `r add_gloss("INT")` — intensifier; `r add_gloss("IV")` — class IV; `r add_gloss("J")` — class IV; `r add_gloss("LAT")` — lative; `r add_gloss("LNK")` — linking suffix; `r add_gloss("M")` — masculine class; `r add_gloss("N")` — neuter class; `r add_gloss("N2")` — second neuter class; `r add_gloss("NUM")` — numeral marker; `r add_gloss("OBL")` — oblique; `r add_gloss("ORD")` — ordinal; `r add_gloss("PFV")` — perfective; `r add_gloss("SG")` — singular

## References
