```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = '')
library(lingglosses)
```

## Introduction

This chapter is dedicated to the origins of numeral markers across the East Caucasian family and neighboring languages.

## Results

Numeral markers in East Caucasian seem to be semantically vacuous, but in some languages they coincide with affixes of different categories. The first map shows the distribution of known origins of numeral markers in the languages of the family. In the Avar-Andic branch and some other languages, numeral markers coincide with intensifying particles [@forker2015]. These particles may originate in copulas ‘to be’ [@maisakchumakina2001: 718]. Consider the following examples from Standard Avar which show different uses of this suffix:

(@) Standard Avar [@forker2020: 13]
```{r, echo=FALSE}
gloss_example(transliteration = "anc’-**go**",
              glosses = "ten-**NUM**",
              free_translation = "10"
)
```

(@) Standard Avar [@forker2015: 8]
```{r, echo=FALSE}
gloss_example(transliteration = "do-b=**go** ħami-de-gun",
              glosses = "DEM-N=**INT** donkey.OBL-LAT-COM",
              free_translation = "the very same donkey"
)
```

In Lezgic languages numerals usually attach a class marker which agrees with the corresponding noun, as exemplified in (@rutul_three_sisters). In Lezgian and Agul, nominal class is lost, and former class markers are petrified as usual suffixes.

(@rutul_three_sisters) Kina Rutul [@mukhin2020: 2]
```{r, echo=FALSE}
gloss_example(transliteration = "xibɨ-**r** riši",
              glosses = "three-**II** sister.II",
              free_translation = "three sisters"
)
```

In Lak and Archi numeral markers coincide with emphatic particles, which have a class slot inside. In the following examples the same particle has different functions:

(@) Archi [@mikailov1967: 75]
```{r, echo=FALSE}
gloss_example(transliteration = "wic’-**i<t’>u**",
              glosses = "ten-**NUM<IV.SG>**",
              free_translation = "10"
)
```

(@) Archi [@chumakina2021: 456]
```{r, echo=FALSE}
gloss_example(transliteration = "jamu-tː=**ij<t’>u** ekutːut nokɬ’=u",
              glosses = "this-IV.SG=**<IV.SG>EMPH** fall.PFV.ATTR.IV.SG house=and",
              free_translation = "the same wrecked (falling apart) house",
              line_length = 100
)
```

In Akhvakh of the Andic branch and Tsakhur of the Lezgic branch cardinals use more than one suffix. In Akhvakh the suffix -*da* is followed by the suffix -*CL-e*. -*da* is also the suffix of adjectives, possibly originating in a participle *ida-CL-e* ‘being’ [@magomedbekova1967: 67]:

(@) Northern Akhvakh [@magomedbekova1967: 71]
```{r, echo=FALSE}
gloss_example(transliteration = "ĩƛːi-**da-be**",
              glosses = "six-**NUM-CM**",
              free_translation = "6"
)
```


In Tsakhur, on the other hand, the class marker is followed by an additional suffix -*le*. Functions of this morpheme are ‘similar to those of an attributivizer’ [@kibrik1999: 158]:

(@) Mishlesh Tsakhur [@kibrik1999: 158]
```{r, echo=FALSE}
gloss_example(transliteration = "q’oˤ=**b-le**",
              glosses = "two=**CM-NUM**",
              free_translation = "2"
)
```

In most Tsezic languages numerals end with -*no*/-*na* (-*ra* in oblique cases), whose origin is hard to trace. Curiously, according to grammars, this marker coincides with the bisyndetic conjunctive suffix which is used as a linking suffix in complex numerals. The numeral marker -*no*/-*na* and the conjunctive suffix coincide in four of five Tsezic languages. However, these morphemes differ in behavior. The distinction between the two and the possible diachronic relationship between them has not been studied yet, and in specific cases can appear vague. Consider example (@hinuq_53) from Hinuq of the Tsezic branch:

(@hinuq_53) Hinuq [@forker2013: 394]
```{r, echo=FALSE}
gloss_example(transliteration = "q’o-**no** qu=no oc’e-**no** ɬo-**no**",
              glosses = "two-**NUM** twenty=ADD ten-**NUM** three-**NUM**",
              free_translation = "53"
)
```

Many Dargic languages have a common suffix -*al*, which is attached by different rules in different lects. Example (@tanty_10) from Tanty shows a typical Dargic cardinal. Its origin is unknown. However, it is worth mentioning that according to Gasanova [-@gasanova1971: 73], -*al* is one of the suffixes typical for adjectives in Dargic languages.

(@tanty_10) Tanty Dargwa [@sumbatovalander2014: 94]
```{r, echo=FALSE}
gloss_example(transliteration = "wec’-**al**",
              glosses = "ten-**NUM**",
              free_translation = "10"
)
```

Finally, some East Caucasian languages have no markers at all. In Ingush, cardinals ‘1’, ‘2’, ‘3’, and ‘5’ attach a glottal stop when headless [@nichols2011: 198]. This glottal stop does not appear in adnominal contexts and thus is also not considered in this chapter. In other numerals there is no marker, as in the Standard Ingush cardinal *it:* ‘10’ [@nichols2011: 210].

Numeral markers also occur in some other languages of the Caucasus, e.g. Abaza [@genko1955]. A similar but not identical task of mapping numeral markers in the languages of the Caucasus has been undertaken in [@moroz2012]. However, I am not aware of any large-scale typological works on the matter.

## Distribution
Lezgic languages generally use class markers as numeral markers and Avar-Andic languages particles with emphatic functions (-*gu*, -*go*, -*da*). Dargic and Tsezic branches use different markers whose origins are not identified. Nakh languages and Khinalug do not have numeral markers.

## List of glosses {-}

```{r}
make_gloss_list(readxl::read_xlsx("../glosses.xlsx"))
```

## References {-}
