

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning=FALSE, fig.width = 9.5)
library(tidyverse)
library(lingtypology)
```

## {-}

Visualization 1 (**General datapoints**) and 3 (**Extrapolated data**) are both based on the principle 'one value – one language'. Visualization 2 (**Data granularity**) takes into account dialect levels. On the **Data granularity** maps you can see the village and the data type (village data, general language data, etc.) when you click on a dot. Hover over or click on a dot to see the language. By unticking the box “show languages” you can remove the inner dots and visualize the distribution of different values in the area without the distraction of genealogical information.

```{r}

feature_dataset <- read_tsv("../orig_table/optative.tsv", show_col_types = FALSE)

feature_dataset |> 
  filter(map == 'yes') ->
  feature_dataset_4map

villages <- read_tsv('../tald_villages.csv', show_col_types = FALSE) # village coordinates
genlang <- read_tsv('../genlangpoints.csv', show_col_types = FALSE) # general language points

feature_dataset_4map |> 
  filter(type == 'language') ->
  feature_dataset_4map_languages

feature_dataset_4map |> 
  filter(type != 'language') ->
  feature_dataset_4map_rest

genlang |> 
  filter(aff == 'Dargwa',
         lang != 'Kaitag',
         lang != 'Standard Dargwa') ->
  change

if(nrow(change) > 0) {
  villages$gltc_lang[which(villages$village_dialect %in% change$gltc_lang
)] <- change$gltc_lang

  villages$gltc_lang[which(villages$dialect_toplevel == 'Kaitag')] <- 'kajt1238'
}

villages |> 
  select(village, rus_village, lat, lon, gltc_lang, lang, aff, family, standard, dialect_toplevel, dialect_nt1, dialect_nt2, dialect_nt3, village_dialect, lang_col, aff_col) |> 
  pivot_longer(names_to = 'type', values_to = 'idiom', standard:village_dialect) |> 
  filter(!is.na(idiom)) |> 
  mutate(type = case_when(type == 'village_dialect' ~ 'village',
                          TRUE ~ type)) |>
  inner_join(feature_dataset_4map_rest, by = c('type', 'idiom', 'lang')) |> 
  distinct() ->
  all_data_without_languages

villages |> 
  select(village, rus_village, lat, lon, gltc_lang, lang, aff, family, lang_col, aff_col) |>
  filter(lang %in% feature_dataset_4map_languages$lang) |>
  anti_join(all_data_without_languages |>  select(village)) |> 
  inner_join(feature_dataset_4map_languages) |> 
  distinct() |> 
  bind_rows(all_data_without_languages)  ->
  alldata_clean  

alldata_clean  |> 
  distinct(gltc_lang) |> 
  mutate(lang4map = lang.gltc(gltc_lang),
         display = 'show languages') |> 
  right_join(alldata_clean) |> 
  filter(!is.na(contributor)) |> 
  mutate(type = factor(type, levels = c('language', 'dialect_toplevel', 'dialect_nt1', 'dialect_nt2', 'dialect_nt3', 'village'))) ->
  alldata_clean

#make short

feature_dataset_4map |> 
  filter(genlang_point == 'yes') |> 
  mutate(lang = ifelse(idiom == 'Standard Dargwa', 'Standard Dargwa', lang)) |> 
  left_join(genlang, by = join_by(lang)) |> 
  distinct() |> 
  mutate(lang4map = lang.gltc(gltc_lang),
         display = 'show languages') |> 
  filter(!is.na(contributor)) ->
  all_genpoints
  
rm(all_data_without_languages, change, feature_dataset_4map, feature_dataset_4map_languages, feature_dataset_4map_rest, villages, genlang)
```

## number of dedicated optatives{.tabset .tabset-fade .tabset-pills #m1} 
    
### General datapoints {-}

```{r}
all_genpoints |> 
  filter(!is.na(lang4map),
    !is.na(all_genpoints$value1)) |> 
  add_count(value1) |> 
  mutate(value1 = str_c(value1, ' (', n, ')')) ->
  all_genpoints_filtered

map.feature(all_genpoints_filtered$lang4map,
            latitude = all_genpoints_filtered$lat,
            longitude = all_genpoints_filtered$lon,
            features = all_genpoints_filtered$value1,
            title = all_genpoints_filtered$value1_name[1],
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            legend = TRUE,
            legend.position = 'bottomleft', 
            zoom.control = TRUE,
            width = 8) %>% 
  map.feature(all_genpoints_filtered$lang4map,
              latitude = all_genpoints_filtered$lat,
              longitude = all_genpoints_filtered$lon,
              features = all_genpoints_filtered$lang,
              color = all_genpoints_filtered$lang_col,
              tile = 'Esri.WorldGrayCanvas',
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = all_genpoints_filtered$display)
```

### Extrapolated data {-}

```{r}
alldata_clean |> 
  filter(!is.na(lang4map),
    !is.na(alldata_clean$value1)) ->
  alldata_clean_filtered

map.feature(alldata_clean_filtered$lang4map,
            latitude = alldata_clean_filtered$lat, 
            longitude = alldata_clean_filtered$lon,
            features = alldata_clean_filtered$value1,
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            title = alldata_clean_filtered$value1_name[1],
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = alldata_clean_filtered$lang,
            zoom.control = TRUE,
            width = 8,
            popup = paste(alldata_clean_filtered$village, '|',
                          alldata_clean_filtered$rus_village, '<br>',
                          'data:', alldata_clean_filtered$type)) %>% 
  map.feature(alldata_clean_filtered$lang4map,
              latitude = alldata_clean_filtered$lat,
              longitude = alldata_clean_filtered$lon,
              features = alldata_clean_filtered$lang,
              color = alldata_clean_filtered$lang_col,
              tile = 'Esri.WorldGrayCanvas',
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = alldata_clean_filtered$display,
              popup = paste(alldata_clean_filtered$village, '|',
                            alldata_clean_filtered$rus_village, '<br>',
                            'data:', alldata_clean_filtered$type))
```

### Data granularity {-}

```{r}
map.feature(alldata_clean_filtered$lang4map,
            latitude = alldata_clean_filtered$lat, 
            longitude = alldata_clean_filtered$lon,
            features = alldata_clean_filtered$value1,
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            title = alldata_clean_filtered$value1_name[1],
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = alldata_clean_filtered$lang,
            zoom.control = TRUE,
            control = alldata_clean_filtered$type,
            popup = paste(alldata_clean_filtered$village, '|',
                          alldata_clean_filtered$rus_village, '<br>',
                          'data:', alldata_clean_filtered$type))
```


    
## positive optative stem{.tabset .tabset-fade .tabset-pills #m2} 
    
### General datapoints {-}

```{r}
all_genpoints |> 
  filter(!is.na(lang4map),
    !is.na(all_genpoints$value2)) |> 
  add_count(value2) |> 
  mutate(value2 = str_c(value2, ' (', n, ')')) ->
  all_genpoints_filtered

map.feature(all_genpoints_filtered$lang4map,
            latitude = all_genpoints_filtered$lat,
            longitude = all_genpoints_filtered$lon,
            features = all_genpoints_filtered$value2,
            title = all_genpoints_filtered$value2_name[1],
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            legend = TRUE,
            legend.position = 'bottomleft', 
            zoom.control = TRUE,
            width = 8) %>% 
  map.feature(all_genpoints_filtered$lang4map,
              latitude = all_genpoints_filtered$lat,
              longitude = all_genpoints_filtered$lon,
              features = all_genpoints_filtered$lang,
              color = all_genpoints_filtered$lang_col,
              tile = 'Esri.WorldGrayCanvas',
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = all_genpoints_filtered$display)
```

### Extrapolated data {-}

```{r}
alldata_clean |> 
  filter(!is.na(lang4map),
    !is.na(alldata_clean$value2)) ->
  alldata_clean_filtered

map.feature(alldata_clean_filtered$lang4map,
            latitude = alldata_clean_filtered$lat, 
            longitude = alldata_clean_filtered$lon,
            features = alldata_clean_filtered$value2,
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            title = alldata_clean_filtered$value2_name[1],
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = alldata_clean_filtered$lang,
            zoom.control = TRUE,
            width = 8,
            popup = paste(alldata_clean_filtered$village, '|',
                          alldata_clean_filtered$rus_village, '<br>',
                          'data:', alldata_clean_filtered$type)) %>% 
  map.feature(alldata_clean_filtered$lang4map,
              latitude = alldata_clean_filtered$lat,
              longitude = alldata_clean_filtered$lon,
              features = alldata_clean_filtered$lang,
              color = alldata_clean_filtered$lang_col,
              tile = 'Esri.WorldGrayCanvas',
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = alldata_clean_filtered$display,
              popup = paste(alldata_clean_filtered$village, '|',
                            alldata_clean_filtered$rus_village, '<br>',
                            'data:', alldata_clean_filtered$type))
```

### Data granularity {-}

```{r}
map.feature(alldata_clean_filtered$lang4map,
            latitude = alldata_clean_filtered$lat, 
            longitude = alldata_clean_filtered$lon,
            features = alldata_clean_filtered$value2,
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            title = alldata_clean_filtered$value2_name[1],
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = alldata_clean_filtered$lang,
            zoom.control = TRUE,
            control = alldata_clean_filtered$type,
            popup = paste(alldata_clean_filtered$village, '|',
                          alldata_clean_filtered$rus_village, '<br>',
                          'data:', alldata_clean_filtered$type))
```


    
## positive optative suffix{.tabset .tabset-fade .tabset-pills #m3} 
    
### General datapoints {-}

```{r}
all_genpoints |> 
  filter(!is.na(lang4map),
    !is.na(all_genpoints$value3)) |> 
  add_count(value3) |> 
  mutate(value3 = str_c(value3, ' (', n, ')')) ->
  all_genpoints_filtered

map.feature(all_genpoints_filtered$lang4map,
            latitude = all_genpoints_filtered$lat,
            longitude = all_genpoints_filtered$lon,
            features = all_genpoints_filtered$value3,
            title = all_genpoints_filtered$value3_name[1],
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            legend = TRUE,
            legend.position = 'bottomleft', 
            zoom.control = TRUE,
            width = 8) %>% 
  map.feature(all_genpoints_filtered$lang4map,
              latitude = all_genpoints_filtered$lat,
              longitude = all_genpoints_filtered$lon,
              features = all_genpoints_filtered$lang,
              color = all_genpoints_filtered$lang_col,
              tile = 'Esri.WorldGrayCanvas',
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = all_genpoints_filtered$display)
```

### Extrapolated data {-}

```{r}
alldata_clean |> 
  filter(!is.na(lang4map),
    !is.na(alldata_clean$value3)) ->
  alldata_clean_filtered

map.feature(alldata_clean_filtered$lang4map,
            latitude = alldata_clean_filtered$lat, 
            longitude = alldata_clean_filtered$lon,
            features = alldata_clean_filtered$value3,
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            title = alldata_clean_filtered$value3_name[1],
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = alldata_clean_filtered$lang,
            zoom.control = TRUE,
            width = 8,
            popup = paste(alldata_clean_filtered$village, '|',
                          alldata_clean_filtered$rus_village, '<br>',
                          'data:', alldata_clean_filtered$type)) %>% 
  map.feature(alldata_clean_filtered$lang4map,
              latitude = alldata_clean_filtered$lat,
              longitude = alldata_clean_filtered$lon,
              features = alldata_clean_filtered$lang,
              color = alldata_clean_filtered$lang_col,
              tile = 'Esri.WorldGrayCanvas',
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = alldata_clean_filtered$display,
              popup = paste(alldata_clean_filtered$village, '|',
                            alldata_clean_filtered$rus_village, '<br>',
                            'data:', alldata_clean_filtered$type))
```

### Data granularity {-}

```{r}
map.feature(alldata_clean_filtered$lang4map,
            latitude = alldata_clean_filtered$lat, 
            longitude = alldata_clean_filtered$lon,
            features = alldata_clean_filtered$value3,
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            title = alldata_clean_filtered$value3_name[1],
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = alldata_clean_filtered$lang,
            zoom.control = TRUE,
            control = alldata_clean_filtered$type,
            popup = paste(alldata_clean_filtered$village, '|',
                          alldata_clean_filtered$rus_village, '<br>',
                          'data:', alldata_clean_filtered$type))
```


    
## negative optative stem{.tabset .tabset-fade .tabset-pills #m4} 
    
### General datapoints {-}

```{r}
all_genpoints |> 
  filter(!is.na(lang4map),
    !is.na(all_genpoints$value4)) |> 
  add_count(value4) |> 
  mutate(value4 = str_c(value4, ' (', n, ')')) ->
  all_genpoints_filtered

map.feature(all_genpoints_filtered$lang4map,
            latitude = all_genpoints_filtered$lat,
            longitude = all_genpoints_filtered$lon,
            features = all_genpoints_filtered$value4,
            title = all_genpoints_filtered$value4_name[1],
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            legend = TRUE,
            legend.position = 'bottomleft', 
            zoom.control = TRUE,
            width = 8) %>% 
  map.feature(all_genpoints_filtered$lang4map,
              latitude = all_genpoints_filtered$lat,
              longitude = all_genpoints_filtered$lon,
              features = all_genpoints_filtered$lang,
              color = all_genpoints_filtered$lang_col,
              tile = 'Esri.WorldGrayCanvas',
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = all_genpoints_filtered$display)
```

### Extrapolated data {-}

```{r}
alldata_clean |> 
  filter(!is.na(lang4map),
    !is.na(alldata_clean$value4)) ->
  alldata_clean_filtered

map.feature(alldata_clean_filtered$lang4map,
            latitude = alldata_clean_filtered$lat, 
            longitude = alldata_clean_filtered$lon,
            features = alldata_clean_filtered$value4,
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            title = alldata_clean_filtered$value4_name[1],
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = alldata_clean_filtered$lang,
            zoom.control = TRUE,
            width = 8,
            popup = paste(alldata_clean_filtered$village, '|',
                          alldata_clean_filtered$rus_village, '<br>',
                          'data:', alldata_clean_filtered$type)) %>% 
  map.feature(alldata_clean_filtered$lang4map,
              latitude = alldata_clean_filtered$lat,
              longitude = alldata_clean_filtered$lon,
              features = alldata_clean_filtered$lang,
              color = alldata_clean_filtered$lang_col,
              tile = 'Esri.WorldGrayCanvas',
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = alldata_clean_filtered$display,
              popup = paste(alldata_clean_filtered$village, '|',
                            alldata_clean_filtered$rus_village, '<br>',
                            'data:', alldata_clean_filtered$type))
```

### Data granularity {-}

```{r}
map.feature(alldata_clean_filtered$lang4map,
            latitude = alldata_clean_filtered$lat, 
            longitude = alldata_clean_filtered$lon,
            features = alldata_clean_filtered$value4,
            color = 'magma',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            title = alldata_clean_filtered$value4_name[1],
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = alldata_clean_filtered$lang,
            zoom.control = TRUE,
            control = alldata_clean_filtered$type,
            popup = paste(alldata_clean_filtered$village, '|',
                          alldata_clean_filtered$rus_village, '<br>',
                          'data:', alldata_clean_filtered$type))
```


    


## Datatable

[Download](https://raw.githubusercontent.com/LingConLab/TALD/master/data/orig_table/optative.tsv) the whole dataset and [its description](https://raw.githubusercontent.com/LingConLab/TALD/master/data/md_description/optative.md).

```{r}
bib <- RefManageR::ReadBib(file = '../bibliography.bib')

feature_dataset |> 
  select(str_which(colnames(feature_dataset), 'value\\d{1,}_name$')) |> 
  pivot_longer(cols = everything()) |> 
  distinct() |>  
  mutate(name = str_remove(name, '_name')) ->
  columns_rename

feature_dataset |> 
  select(lang, idiom, source, page, matches('value\\d{1,}$')) |> 
  rename_with(function(x){columns_rename$value[match(x, columns_rename$name)]}, matches('value\\d{1,}$'))  |> 
  rename(Language=lang, 
         Idiom = idiom,
         Source = source) |> 
  mutate(page = str_replace_all(page, '--', '–'),
         Source = str_split(Source, '; '),
         page = str_split(page, '; ')) |> 
  unnest_longer(col = c(Source, page)) |> 
  rowwise() |> 
  mutate(page = ifelse(page == 'NA', NA_character_, page),
         Source = ifelse(str_detect(Source, 'p.c.$'),
                         Source,
                         Cite(bib, Source, 
                       after = ifelse(!is.na(page), 
                                      str_c(': ', page),
                                      '')))) |>
  select(-page) |> 
  DT::datatable(class = 'cell-border stripe', 
    rownames = FALSE, 
    filter = 'top', 
    extensions = 'Buttons',
    options = list(pageLength = 100, 
                   autoWidth = TRUE, 
                   info = FALSE,
                   dom = 'fBltp',
                   buttons = list(list(extend = 'collection',
                                       buttons = c('csv', 'excel', 'pdf'),
                                       text = '<i class="fas fa-download"></i>')),
                   paginate = TRUE))
```

## References {-}


