

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning=FALSE, fig.width = 9.5)
library(tidyverse)
library(lingtypology)
library(RCaucTile)
```

## {-}

Visualization 1 (**General datapoints**) shows no more than one dot per language. Visualization 2 (**Extrapolated data**) represents each language as a cluster of dots, which correspond to villages where a certain language is spoken. Visualization 3 (**Data granularity**) takes into account dialect levels. On the **Data granularity** maps you can see the village and the data type (village data, general language data, etc.) when you click on a dot. Hover over or click on a dot to see the language. By unticking the box “show languages” you can remove the inner dots and visualize the distribution of different values in the area without the distraction of genealogical information.

```{r}

read_tsv("../orig_table/ordinals.tsv", show_col_types = FALSE, guess_max = 2000) |>
  mutate(type = case_when(type == 'language' ~ 'language',
                          type == 'dialect_toplevel' ~ 'top level dialect',
                          type == 'dialect_nt1' ~ 'non top level 1 dialect',
                          type == 'dialect_nt2' ~ 'non top level 2 dialect',
                          type == 'dialect_nt3' ~ 'non top level 3 dialect',
                          type == 'village' ~ 'village dialect')) |> 
  filter(map != 'no') ->
  feature_dataset

read_tsv('../tald_villages.csv', show_col_types = FALSE, guess_max = 2000)  |>
  select(village, rus_village, lat, lon, gltc_lang, aff, family, standard, default_level, dialect_toplevel, dialect_nt1, dialect_nt2, dialect_nt3, village_dialect, lang_col, aff_col) |> 
  mutate(language = default_level) |>
  pivot_longer(names_to = 'type', values_to = 'idiom', standard:village_dialect) |> 
  filter(!is.na(idiom)) |> 
  mutate(type = case_when(type == 'standard' ~ 'language',
                          type == 'default_level' ~ 'language',
                          type == 'dialect_toplevel' ~ 'top level dialect',
                          type == 'dialect_nt1' ~ 'non top level 1 dialect',
                          type == 'dialect_nt2' ~ 'non top level 2 dialect',
                          type == 'dialect_nt3' ~ 'non top level 3 dialect',
                          type == 'village_dialect' ~ 'village dialect')) ->
  coordinates

coordinates |> 
  distinct(gltc_lang) |> 
  mutate(lang4map = lang.gltc(gltc_lang)) ->
  lang4map

coordinates |> 
  group_by(idiom, type, gltc_lang, lang_col) |> 
  reframe(lat = mean(lat), 
          lon = mean(lon)) |> 
  left_join(lang4map)->
  coordinates_averaged

feature_dataset |> 
  mutate(idiom = str_remove(idiom, 'Standard ')) |> 
  inner_join(coordinates, by = c('type', 'idiom', 'language'), relationship = 'many-to-many') |> 
  left_join(lang4map) |> 
  mutate(display = 'show languages') ->
  all_data
```

## 'say' strategy {.tabset .tabset-fade .tabset-pills #m1} 
    
### General datapoints {-}

```{r}
feature_dataset |> 
  filter(map == 'yes',
         genlang_point == 'yes') |>
  add_count(value1) |> 
  inner_join(coordinates_averaged) |> 
  mutate(popup = ifelse((lang4map == idiom | str_detect(idiom, 'Standard')), 
                        str_c('data level: ', type),
                        str_c(idiom, '<br> data level: ', type)),
         value1 = str_c(value1, ' (', n, ')'),
         display = 'show languages')  |> 
  filter(!is.na(value1),
         !is.na(lang4map)) ->
  general_datapoints_map  

map.feature(general_datapoints_map$lang4map,
            latitude = general_datapoints_map$lat,
            longitude = general_datapoints_map$lon,
            label = general_datapoints_map$language,
            features = general_datapoints_map$value1,
            color = 'viridis',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            legend = TRUE,
            legend.position = 'bottomleft', 
            zoom.control = TRUE,
            width = 8) %>% 
  map.feature(general_datapoints_map$lang4map,
              latitude = general_datapoints_map$lat,
              longitude = general_datapoints_map$lon,
              features = general_datapoints_map$language,
              label = general_datapoints_map$language,
              color = general_datapoints_map$lang_col,
              tile = 'Esri.WorldGrayCanvas',
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = general_datapoints_map$display)
```

### Extrapolated data {-}

```{r}
all_data |> 
  filter(!is.na(value1)) ->
  all_data_filtered

map.feature(all_data_filtered$lang4map,
            latitude = all_data_filtered$lat, 
            longitude = all_data_filtered$lon,
            features = all_data_filtered$value1,
            color = 'viridis',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = all_data_filtered$language,
            zoom.control = TRUE,
            width = 8,
            popup = paste(all_data_filtered$village, '|',
                          all_data_filtered$rus_village, '<br>',
                          'data:', all_data_filtered$type)) %>% 
  map.feature(all_data_filtered$lang4map,
              latitude = all_data_filtered$lat,
              longitude = all_data_filtered$lon,
              features = all_data_filtered$language,
              label = all_data_filtered$language,
              color = all_data_filtered$lang_col,
              tile = 'Esri.WorldGrayCanvas',
              legend = FALSE,
              width = 5,
              pipe.data = .,
              control = all_data_filtered$display,
              popup = paste(all_data_filtered$village, '|',
                            all_data_filtered$rus_village, '<br>',
                            'data:', all_data_filtered$type))
```

### Data granularity {-}

```{r}
feature_dataset |> 
  filter(map == 'yes') |> 
  inner_join(coordinates_averaged) |> 
  mutate(popup = ifelse((lang4map == idiom | str_detect(idiom, 'Standard')), 
                        str_c('data level: ', type),
                        str_c(idiom, '<br> data level: ', type)),
         type = factor(type, levels = c('language', 
                                        'top level dialect',
                                        'non top level 1 dialect',
                                        'non top level 2 dialect',
                                        'non top level 3 dialect',
                                        'village dialect')))  |> 
  filter(!is.na(value1),
         !is.na(lang4map)) |> 
  arrange(type) ->
  data_granularity_map
  
map.feature(data_granularity_map$lang4map,
            latitude = data_granularity_map$lat, 
            longitude = data_granularity_map$lon,
            features = data_granularity_map$value1,
            color = 'viridis',
            stroke.features = 'a',
            stroke.color = 'black',
            stroke.radius = 1,
            stroke.legend = FALSE,
            tile = 'Esri.WorldGrayCanvas',
            legend = TRUE,
            legend.position = 'bottomleft', 
            label = data_granularity_map$language,
            zoom.control = TRUE,
            control = data_granularity_map$type,
            popup = data_granularity_map$popup)
```

### Tile map {-}

```{r, fig.width=12, fig.height=8}
general_datapoints_map |> 
  select(language, value1) |> 
  ec_tile_map(feature_column = 'value1',
              tile_colors = 'viridis') +
  theme(text = element_text(size = 15))
```

    


## Datatable

[Download](https://raw.githubusercontent.com/LingConLab/TALD/master/data/orig_table/ordinals.tsv) the whole dataset.

```{r}
bib <- RefManageR::ReadBib(file = '../bibliography.bib')

feature_dataset |> 
  select(str_which(colnames(feature_dataset), 'feature\\d{1,}$')) |> 
  pivot_longer(cols = everything()) |> 
  distinct() |>  
  mutate(name = str_replace(name, 'feature', 'value')) ->
  columns_rename

feature_dataset |> 
  select(language, idiom, source, page, matches('value\\d{1,}$')) |> 
  filter(!if_all(matches('value\\d{1,}$'), is.na)) |> 
  rename_with(function(x){columns_rename$value[match(x, columns_rename$name)]}, matches('value\\d{1,}$'))  |> 
  rename(Language=language, 
         Idiom = idiom,
         Source = source) |> 
  mutate(page = str_replace_all(page, '--', '–'),
         Source = str_split(Source, '; '),
         page = str_split(page, '; ')) |> 
  unnest_longer(col = c(Source, page)) |> 
  rowwise() |> 
  mutate(page = ifelse(page == 'NA', NA_character_, page),
         Source = case_when(str_detect(Source, '[Ff]ield [Dd]ata') ~ Source,
                            str_detect(Source, 'p.\\s?c.$') ~ Source,
                            TRUE ~ Cite(bib, Source, 
                                        after = ifelse(!is.na(page),  str_c(': ', page), '')))) |> 
  select(-page) |> 
  DT::datatable(class = 'cell-border stripe', 
    rownames = FALSE, 
    filter = 'top', 
    extensions = 'Buttons',
    options = list(pageLength = 100, 
                   autoWidth = TRUE, 
                   info = FALSE,
                   dom = 'fBltp',
                   buttons = list(list(extend = 'collection',
                                       buttons = c('csv', 'excel', 'pdf'),
                                       text = '<i class="fas fa-download"></i>')),
                   paginate = TRUE))
```

## References {-}


