```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = '')
library(lingglosses)
```

## Introduction

In the tradition of linguistic description of the languages of the Caucasus, _translative_ is a term used to refer to a locative case or an adposition with the meaning of motion _through_ a spatial domain. In English, prepositions such as _over_, _through_, _across_ and _along_ would fit the definition of _translative prepositions_ [@comriepolinsky1998: 99].

A considerable number of Nakh-Daghestanian languages, which make up most of the languages in my sample, have a _bimorphemic_ system in which locative “cases” combine two different case suffixes: a marker of _localization_, which defines the spatial domain and includes meanings like ‘in’, ‘at’, ‘under’, ‘on’, etc., and a _directionality_ marker which indicates the type of location or motion, i.e. ‘to’, ‘from’, ‘through’, as well as absence of motion. One example of such a system is the case system of Khwarshi:

##### **Table 1.** Khwarshi (Khalilova, 2009: 74) {#t1 -}
 
|       | Essive | Lative | Versative | Ablative | Translative | Terminative |
|:-----:|:------:|:------:|:---------:|:--------:|:-----------:|:-----------:|
| SUPER | -ƛ’o   | -ƛ’o-l | -ƛ’o-ɣul  | -ƛ’o-zi  | -ƛ’o-ɣužaz  | -ƛ’o-q’a    |
| SUB   | -ƛʰ    | -ƛʰ-ul | -ƛʰ-ɣul   | -ƛʰ-zi   | -ƛʰ-ɣužaz   | -ƛʰ-q’a     |
| IN    | -ma    | -ma-l  | -ma-ɣul   | -ma-zi   | -ma-ɣužaz   | -ma-q’a     |
| INTER | -ɬ     | -ɬ-ul  | -ɬ-ɣul    | -ɬ-zi    | -ɬ-ɣužaz    | -ɬ-q’a      |
| AD    | -ho    | -ho-l  | -ho-ɣul   | -ho-zi   | -ho-ɣužaz   | -ho-q’a     |
| APUD  | -ɣo    | -ɣo-l  | -ɣo-ɣul   | -ɣo-zi   | -ɣo-ɣužaz   | -ɣo-q’a     |
| CONT  | -qo    | -qo-l  | -qo-ɣul   | -qo-zi   | -qo-ɣužaz   | -q-q’a      |

Additionally, there are languages featuring trimorphemic systems in which the third morpheme has deictic, orientational or approximative meanings [@nogina2023]. However, for this chapter the difference between bimorphemic and trimorphemic systems is irrelevant, as it does not impact path-encoding.

In this work, directionality markers with translative semantics were classified as **dedicated translative cases**. These markers are used specifically and only to mark the translative meaning of motion ‘through’ the spatial domain. Such dedicated markers are found both in languages with bimorphemic systems (cf. [Table 1](#t1)). and in languages with monomorphemic systems, such as Chechen:

(@one) Standard Chechen [@molochieva2010: 210]

```{r, results='asis'}
gloss_example(transliteration = "so kuoraa-**xula** gu-sh j-ara cu shimmuo hu liela-d-o",
              glosses = "1SG.NOM(J) window-**TRANS**  see:IPFV-CVBsim J-be.PST DEM two.ERG what carry:INF-D-make:IPFV.PRS",
              free_translation = "I (could) see through the window what those two were doing.")
```

Some languages do not have a dedicated translative marker and use a **non-dedicated suffix**, i.e. a non-translative marker with a secondary translative meaning.

(@two) Tabasaran [@babaliyeva2013: 39]

```{r}
gloss_example(transliteration = "žærɣlar.i-**ʔan**-sina a<b>qh-nu",
              glosses = "rock.PL-**INEL**-down <N>fall-AOR",
              free_translation = "He fell down through the rocks.")
```

In (@two) an elative directionality marker is used. Usually, elative markers indicate motion ‘from’ (out of) the spatial domain, yet in this context the marker bears translative meaning.

If a language does not express translative meaning by case marking alone, this function is usually fulfilled by **adpositions**:

(@three) Tat: Judeo-Tat [@authier2012: 54]

```{r}
gloss_example(transliteration = "u-ho edembir-yt jes-de **ez** tepei",
              glosses = "3-PL FOC.PST-3PL jump-PART **ABL** bonfire",
              free_translation = "They were jumping through the bonfire.")
```

(@four) Tindi [@magomedova2003: 45]

```{r}
gloss_example(transliteration = "ɬesːa-**qa** **aqa**",
              glosses = "river-**ALL** **along**",
              free_translation = "Along the river.")
```

It is important to note that many languages use a combination of both case marking and adposition, compare (@three) and (@four). Yet in my classification they are still classified as marking translative meaning with an adposition if the translative meaning cannot be expressed without it.

## Results

Among the sampled languages non-dedicated translative marking appears to be the prevailing strategy (18 out of 40 languages and dialects sampled), although non-dedicated markers are also common (17 out of 40 languages). Thus, case marking is the dominant strategy for the expression of translative meaning in Daghestan. If we only consider the members of the Nakh-Daghestanian language family, to which the vast majority of the languages in the sample belong, non-dedicated case marking and dedicated case marking are found in 16 languages each. Thus, the expression of translative meaning through case marking is found in 32 out of 34 Nakh-Daghestanian languages in the area. 

(@five) Udi: Nizh [@ganenkov2008: 41]

```{r}
gloss_example(transliteration = "me durb-in-**aχun** χe=ne cːoroj-e-sa.",
              glosses = "PROX pipe-O-**ABL** water=3SG flow-LV-PRS",
              free_translation = "Water flows through this pipe.")
```

(@six) Tsakhur: Mishlesh [@kibrik1999: 174]

```{r, results='asis'}
gloss_example(transliteration = "gade-b-iš-e q'ol-n-ɢ-ē-q'ol-n-ɢ-ē alʁah-a-jī jilʁ-ī-l-**e** dawar-ā-r",
              glosses = "boy-PL-OBL.PL-ERG two-A-OBL.1-ERG-two-A-OBL.1-ERG NPL.move-IPF-EM1 bridge-OBL-SUP-**EL** ram-PL-NOM.PL",
              free_translation = "Boys moved rams across the bridge in pairs.")
```

Almost all non-dedicated translative markers are either ablative (@five) or elative (@six) suffixes. The labels _ablative_ and _elative_ are both used to indicate motion _from_. Different labels are thus the result of different terminological choices, and do not seem to imply any semantic difference in terms of directional properties. 

The only exception to this general rule is Armenian, which uses the instrumental case to mark translative meanings:

(@seven) Armenian [@dum-tragut2009: 91]

```{r}
gloss_example(transliteration = "Get-**ov** anc’n-el t’e kamurj-**ov** gn-al?",
              glosses = "river-**INST** pass-INF or bridge-**INST** go-INF",
              free_translation = "(Shall we) pass the river or go over the bridge?")
```

Interestingly, the same rule does not apply to locative cases combined with adpositions. Languages use a variety of cases, including allative (@four), ablative (@eight) or no case at all (@three).

(@eight) Nogai [@baskakov1973: 276]

```{r}
gloss_example(transliteration = "keš sajɯn kullɯk-tan šɯk-sa ol sosɯ kœpir-**den** œtedi",
              glosses = "evening every work-ABL leave-PST 3SG DET bridge-**ABL** through",
              free_translation = "Every evening after works he leaves through this bridge.")
```

Sometimes a given language can express translative meanings in two different non-dedicated ways.

(@nine) Agul [@magometov1970: 83]

  a.
```{r}
gloss_example(transliteration = "calik-**di**",
              glosses = "wall-**DIR**",
              free_translation = "Along a wall")
```
  b.
```{r}
gloss_example(transliteration = "muxurik-**di**",
              glosses = "chest-**DIR**",
              free_translation = "Along a (human) chest")
```
  c.
```{r}
gloss_example(transliteration = "jux xulaq-**di**",
              glosses = "go.IMP house-**DIR**",
              free_translation = "Go in the direction of my house!")
```

Agul has a directive case that can express translative meanings but only in combination with certain stems that have the topological properties of a vertical surface (compare @nine a, @nine b and @nine c). In other cases, Agul speakers resort to elative marking:

(@ten) Agul [@maisak2014: 217]

```{r}
gloss_example(transliteration = "zun ruq’-ar.i-k-**es** kejč’w.a-jdi qːadaq e p.u-ne",
              glosses = "1SG iron-PL-CONT-**ELAT** {CONT}get.out.IPF-PT:HAB nail COP speak.PF-AOR",
              free_translation = "'I am the nail, that goes through the iron' – he said.")
```

In some languages, the presence of a dedicated translative marker is dubious, as the marker might have been recorded in the past, but is now being ousted by some other case marker, usually by the elative. In Bagvalal, “[a]ccording to informants, this marker [translative] often can be omitted without making the sentence ungrammatical” [@daniel2001: 143], with no further details provided. In Botlikh we observe a different situation, as the original translative marker is now used with elative meanings too, and is partially ousting the original elative marker [@alexeyevverheesdraft2020]. 

(@eleven) Georgian [@hewitt1995: 149]

```{r}
gloss_example(transliteration = "**gada**-di-s",
              glosses = "**THROUGH**-go-3SG.SUBJ",
              free_translation = "X is going over/across/through")
```

The only language in the sample that employs a different strategy is Georgian, which expresses translative meanings with a **preverb**. Some Nakh-Daghestanian languages can also do so derivationally, but unlike them the Georgian translative preverb is not derivational.

(@twelve) Georgian [@hewitt1995: 151, 189]

a.
```{r}
gloss_example(transliteration = "pot-**sa** da odesa-**s** šoris gem-eb-i **da**-di-an",
              glosses = "Poti-**DAT** and Odessa-**DAT** between boat-PL-NOM **GO.REGULARLY**-go-PL",
              free_translation = "Boats (regularly) journey between Poti and Odessa.")
```
b.
```{r}
gloss_example(transliteration = "opl-ma axalux-**s** **ga**-a-t’an-a",
              glosses = "sweat-ERG shirt-**DAT** **THROUGH**-3SG.LV-penetrate-3SG",
              free_translation = "Sweat penetrated through the shirt.")
```

It is also certain that the Georgian translative is not distinctly marked on the dependent, as the dependent only receives dative case-marking that can be applied to a variety of locative meanings. Compare (@twelve a), where the destinations are marked by dative and (@twelve b), where the medium is marked by dative. In both examples the dependents are marked the same, while the verbs have different preverbs that express different locative relations. It can be thus concluded that the actual meaning of translativity in Georgian is conveyed primarily by the preverb and not by case-marking. This makes Georgian a unique case among the languages sampled, so it deserves a separate classification.

## Distribution

Regarding Nakh-Daghestanian languages, there is no clear genealogical relationship that can be drawn for using either a dedicated case, a non-dedicated case, or an adposition.

Avar has a dedicated translative case. Within the Andic branch, Akhvakh, Bagvalal, Botlikh and Godoberi use a dedicated case, while Andi, Chamalal and Karata use ablative/elative. All Tsezic languages use a dedicated marker, except for Hinuq.

Mehweb, Tanty and Kubachi Dargwa all have a dedicated translative, while Standard Dargwa and Itsari do not. 
All sampled Lezgic languages have non-dedicated translative markers, except for Archi. 

The presence of a dedicated marker in Archi could be explained areally, as Archi is on the outskirts of what can be considered a general Lezgic area and is in close contact with Lak as well as Avar. These languages have a dedicated translative case, which might have influenced Archi.
	
As for non-Nakh-Daghestanian languages, Armenian is the only language using instrumental. Kumyk is the only Turkic language in the area that uses a non-dedicated locative case (i.e. ablative) to mark translative meanings.

## List of glosses {-}

```{r}
make_gloss_list(readxl::read_xlsx("../glosses.xlsx"))
```

## References {-}
