```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = '')
library(lingglosses)
```

## Introduction

The imperative is a grammatical mood that expresses a command. The prohibitive is a negative imperative, a form denoting prohibition. This chapter focuses on the ways plural imperatives and prohibitives are marked in the languages of Daghestan, and primarily on whether or not these plural morphemes are dedicated to commands and how they interact with transitivity of the verb. 

According to WALS, out of all the languages that have a morphological imperative, most have distinct morphological forms for the second person singular and the second person plural (79 percent in [@vanderauweraetal2013]). This means that cross-linguistically there is a tendency for languages to distinguish between singular and plural imperatives, if they have the imperative at all. How does Daghestan as a linguistic area typologically compare to the rest of the world? According to the analysis presented in this chapter, Daghestan is indeed distinct in this regard, but not dramatically: 60 percent of the languages sampled distinguish between singular and plural imperatives. This study also explores plural marking on prohibitives, which WALS does not. 

Expression of the plural on commands should be distinguished from gender-number marking on verbs, cf. (@one a,b) and (@one c,d), which is present in most Nakh-Daghestanian languages [@zakirova2023: 74].

(@one) Rutul: Mukhad [@maxmudova2002: 159, 183]
  a.
```{r}
gloss_example(transliteration = "j-ɨq'-a",
              glosses = "M.SG-come-IMP",
              free_translation = "Come! (addressing a singular speaker)")
```
  b.
```{r}
gloss_example(transliteration = "d-ɨq'-a",
              glosses = "HPL-come-IMP",
              free_translation = "Come! (addressing several speakers)")
```
  c.
```{r}
gloss_example(transliteration = "j-ɨq'-as",
              glosses = "M.SG-come-INF",
              free_translation = "to come (singular)")
```
  d.
```{r}
gloss_example(transliteration = "d-ɨq'-as",
              glosses = "HPL-come-INF",
              free_translation = "to come (plural)")
```

Some Nakh-Daghestanian languages may additionally express number on the verb by (irregular) means independent from gender (@two).

(@two) Ingush [@nichols2011: 313], as cited by [@zakirova2023: 103]
  a.
```{r}
gloss_example(transliteration = "ɨz wa-xei**ra**",
              glosses = "3SG down-sit.**WIT**",
              free_translation = "He sat down.")
```
  b.
```{r}
gloss_example(transliteration = "ɨz wa-xei**šar**",
              glosses = "3PL down-sit.**PL.WIT**",
              free_translation = "They sat down.")
```

Number can also be unmarked, as in Lezgian:

(@three) Lezgian [@haspelmath1993: 249, 461]
  a.
```{r}
gloss_example(transliteration = "aqːwaz, ja juldoš-ar za-qʰ jab aka-l*",
              glosses = "Stop(IMP) PT comrade-PL 1SG-POESS ear attach-IMP",
              free_translation = "Stop, comrades, listen to me!")
```
  b.
```{r}
gloss_example(transliteration = "aqːwaz, Peri!",
              glosses = "Stop(IMP) Peri",
              free_translation = "Stop, Peri!")
```

This particular study is concerned specifically with suffixal number marking on imperatives and prohibitives, as gender-number prefixes are the same for all forms of the verb (@one). In this sense, if a given language has no number marking on the imperative outside of gender-number prefix, we view this as absence of number marking on the imperative, cf. (@four).

(@four) Bagvalal: Kvanada [@kibrik2001: 321]
```{r}
gloss_example(transliteration = "o-b bišːdi-b-da partal b-aχːa b-ah-aː **b**-e-be-ʁala-di heƛ'i o-šːu-r",
              glosses = "this-N you-GEN.N-PTC stuff N-away N-take-POT.INF **HPL**-come-IMP-PTC-DI say this-OBL.M-ERG",
              free_translation = "Come pick up this stuff of yours – he said.")
```

As opposed to cases in which imperative plural markers are absent (@one, @three, @four), in some languages of the area the imperative plural is morphologically **marked**. This type of imperative is further divided into two subtypes: _**general plural marking**_ (e.g. a clitic in (@five a)) and _**plural marking dedicated to commands**_ (@six). In the first type, a plural marker found elsewhere in the language (@five b) is also used on the imperative. In the second type, the plural imperative has a unique, dedicated plural morpheme.

(@five) Udi: Nizh [@maisak2006: 2-3]
  a.
```{r}
gloss_example(transliteration = "äjč'ä ek**-i-nan** beši k'oya",
              glosses = "tomorrow come.IMP**-IMP-2PL** we.POSS house.DAT",
              free_translation = "Tommorow come to our house!")
```  
  b.
```{r}
gloss_example(transliteration = "vaˤn za q'amiš**-nan**?",
              glosses = "you.PL I.DAT understanding**-2PL**",
              free_translation = "Do you understand me? (Addressing several people)")
``` 

(@six) Dargwa: Mehweb [@dobrushina2019: 123]
```{r}
gloss_example(transliteration = "w-aˤbʡ-a-**na** rasul!",
              glosses = "M-kill.PFV-IMP.TR-**IMP.PL** Rasul",
              free_translation = "You all kill Rasul!")
``` 

Similarly, the prohibitive plural can be marked or unmarked. Marked prohibitive plural can also be classified in a similar way: _**general plural marking**_, such as the allocutive suffix in (@seven) and _**plural marking dedicated to commands**_.

(@seven) Dargwa: Tanty [@sumbatova2014: 164, 143]

  a.
```{r}
gloss_example(transliteration = "če-tːi**-ma-**d-irk-u-tː**-a-ja**!",
              glosses = "ON-behind**-PROH-**2PL-fall.IPF-TH-2**-IMP-ALL.PL**",
              free_translation = "Don't win!")
``` 
  b.
```{r}
gloss_example(transliteration = "čaj či-ž dikː-u-l-da**-ja**?",
              glosses = "tea who-DAT 1/2PLwant-PRS-CVB-2PL**-ALL.PL**",
              free_translation = "Who wants tea?")
``` 

Finally, we considered the relationship between plural marking on imperatives and transitivity, as some languages have morphological number distinction only in intransitive verbs, see [@starodubtseva2024].

(@eight) Karata: Anchiq (Konstantin Filatov, p.c.)

  a.
```{r}
gloss_example(transliteration = "men=al hosːa=ra gah-**a**",
              glosses = "you.SG=ADD thus=EMPH do-**IMP**",
              free_translation = "You(SG) do it too!")
``` 
  b.  
```{r}
gloss_example(transliteration = "bišdi-l=ara gah-**a**",
              glosses = "you.PL-ERG=EMPH do-**IMP**",
              free_translation = "You(PL) do it yourselves!")
```   
  c.
```{r}
gloss_example(transliteration = "ʕedero w-ox-**u**",
              glosses = "immediately M-come-**IMP**",
              free_translation = "Come quick!")
``` 
  d.
```{r}
gloss_example(transliteration = "χːeχːala b-exʷ-**abi** ešq-a-r",
              glosses = "fast HPL-come-**IMP.PL** home-SUP-LAT",
              free_translation = "Quick, come home!")
``` 


## Results

Among the sampled languages, plural marking on imperatives seems to be the prevailing strategy (60 percent of the languages) ([Map 1](http://lingconlab.ru/dagatlas/057_plural_marking_on_imperatives_map.html#m1)).
Notably, some languages have multiple imperative forms with varying degrees of politeness. In Ingush the neutral (default) imperative is unmarked for plurality, while the future imperative has plural marking [@Nichols2011: 274-275]. In cases like this, we decided to map only the neutral imperative form. 
Fewer languages mark the plural prohibitive ([Map 2](http://lingconlab.ru/dagatlas/057_plural_marking_on_imperatives_map.html#m2)). There are no cases in which  imperative plural is absent but prohibitive plural is present. Out of all languages that have plural imperatives, around 60 percent also have plural prohibitives ([Map 5](http://lingconlab.ru/dagatlas/057_plural_marking_on_imperatives_map.html#m5)).

Based on the imperative morpheme types mentioned (*unmarked*, *general plural marking* and *plural marking dedicated to commands*), it was assumed that some languages could also have *dedicated prohibitive plurals*. However, no such morphemes were found among the sampled languages. The prohibitive, if pluralised at all, is pluralised using the imperative plural marker (cf. (@nine) and (@six)).

(@nine) Dargwa: Mehweb [@dobrushina2019: 128]

```{r}
gloss_example(transliteration = "deč' mi-m-iq'**-adi-na**!",
              glosses = "song NEG.VOL-N-do.IPFV-**PROH-IMP.PL**",
              free_translation = "Don't sing! (addressing several people)")
``` 

An exception to this general rule is Chirag Dargwa, where the prohibitive plural (@ten a) and the imperative plural (@ten b) are marked using different morphemes:

(@ten) Dargwa: Chirag [@ganenkov2021draft]
  a.
```{r}
gloss_example(transliteration = "uncːa	me-bšː-i**-tːa**",
              glosses = "door(ABS) PROH-open.IPF-**PROH**.TR-**PL**",
              free_translation = "Don't open the door! (addressing several people)")
```

  b.
```{r}
gloss_example(transliteration = "ca čuja b-ic-a-**ja**",
              glosses = "one ram(ABS) N.SG-sell.PF-**IMP-PL**",
              free_translation = "Sell one ram! (addressing several people)")
```

But even in Chirag prohibitives can be optionally double-marked for plurality with the imperative plural morpheme *-ja*.

Out of the 25 languages that have the plural imperative, only in 5 plural marking is dependent on the transitivity of the verb, with intransitive imperatives being marked for plural [Map 3](http://lingconlab.ru/dagatlas/057_plural_marking_on_imperatives_map.html#m3). Among them is Kubachi Dargwa, which has this split only in one conjugation (@eleven). While transitive imperatives of the 1st conjugation class do not mark plurality (@eleven a, b), intransitive imperatives of the 1st conjugation class do (@eleven c, d). All verbs belonging to the 2nd and 3rd classes distinguish number irrespective of their transitivity.

(@eleven) Dargwa: Kubachi [@magometov1963: 202]

  a.
```{r}
gloss_example(transliteration = "b-iːq'-**a**",
              glosses = "N-do-**IMP**",
              free_translation = "Do!")
```
  b.
```{r}
gloss_example(transliteration = "b-iːq'-**a**",
              glosses = "N-do-**IMP**",
              free_translation = "Do!")
```
  c.
```{r}
gloss_example(transliteration = "w-iːq'-**e**",
              glosses = "M-work-**IMP**",
              free_translation = "Work!")
```
  d.
```{r}
gloss_example(transliteration = "d-iːq'-**aː**",
              glosses = "HPL-work-**IMP.PL**",
              free_translation = "Work!")
```


There are no cases of similar splits in prohibitive forms [Map 4](http://lingconlab.ru/dagatlas/057_plural_marking_on_imperatives_map.html#m4).


## Distribution

Most branches of the Nakh-Daghestanian family show a certain homogeneity when it comes to plural marking on imperatives. All Dargwa and all Nakh languages except Ingush have a morphological plural imperative, while all Tsezic languages lack it. However, there is no clear prevalence of one type over the other among Avar-Andic and Lezgic languages. All languages outside Nakh-Daghestanian uniformly distinguish between singular and plural imperatives. None of them displays plural marking dedicated to commands.

The distribution of the morphological plural prohibitive is less clear. The only connection with the distribution of the plural imperative seems to be that both are marked in all Dargwa languages and unmarked in all Tsezic languages. More generally, it can be observed that languages that lack the plural imperative will certainly lack the prohibitive plural as well. 

The transitivity split in plural marking on imperatives is not very common in general and is only found in some languages of the Avar-Andic, Dargwa and Lezgic branches, while it is completely absent in all the other branches. Notably, the only Lezgic language to show the split is Archi.

## List of glosses {-}

```{r}
make_gloss_list(readxl::read_xlsx("../glosses.xlsx"))
```

## References {-}
