```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = '')
library(lingglosses)
```

## Introduction

Caritive markers denote the non-involvement of a participant in a situation (cf. [@oskolskayaetal2020: 21-22]). Most often the non-involvement predication is an adverbial or attributive expression (@one)-(@two) modifying the situation or a participant of a different situation.

(@one) _to come **without** an invitation_
(@two) _a child**less** woman_

Typologically caritive markers can be more or less dedicated, i.e., the expression of the caritive meaning can be the only function of the given marker or just one of its functions. If a language lacks a dedicated caritive marker, the caritive meaning can be expressed, for example, by a negative possessive construction, cf. (@three)-(@four).

(@three) _to come **not having** an invitation_
(@four) _a woman **not having** children_

Both strategies are attested in the languages of Daghestan: some languages have dedicated caritive markers, cf. the caritive suffix _-t’u_ in Khwarshi (@five), while others express the caritive meaning periphrastically, cf. the participial construction with the existential verb in Agul (@six). Often both strategies are available in one and the same language.

(@five) Khwarshi [@khalilova2009: 263]  
_lok’o_ ‘heart’ > _lok’o-t’u_ ‘heartless’	

(@six) Agul, Huppuq’ dialect [@maisakpanovadraft]
```{r}
gloss_example("baw **qa-dawa** šünükː düj.i-l jaratmiš d-aq’-u-raj",
              "{[}mother **POST.be-NEG(PTCP)**{]} child world-SUP create NEG-do-PFV-JUSS",
              "Let there be no motherless (lit. mother not-having) children in the world!")
```

The present chapter gives an overview of caritive markers attested in the languages spoken in Daghestan and is largely based on [@maisakpanovadraft], where caritive markers in East Caucasian languages are discussed in more detail.


## Results

Two main syntactic classes of caritive phrases were identified: attributive and adverbial.

### Attributive caritive phrases

Within the class of attributive caritive phrases there are two competing strategies for expressing the caritive meaning: a dedicated suffix/clitic and a participle which can be roughly translated as ‘not having (something)’.

The most widespread dedicated attributive caritive marker in the languages of Daghestan is the Turkic suffix _-sız/-siz/-suz/-süz_. It is attested in the Turkic languages of Daghestan (Azerbaijani, Kumyk, Nogai) and in Lezgic languages of the East Caucasian family. This suffix was borrowed from Turkic to Lezgic in the invariable form _-sɨz_ or _-suz_, see, e.g., [@authier2010: 26-27] on the caritive suffix _-suz_ in Kryz. In some Avar-Andic languages we find another group of similar caritive suffixes forming attributive expressions: _-q’-a-b_ in Avar, _-tu-b_ in Bagvalal and Tindi and _-q’u-b_ in Karata (the final slot of all these suffixes is a gender agreement marker). Attributive caritive suffixes were also found in two Tsezic languages (_-taj_ in Tsez and _-t’u_ in Khwarshi), in all three Nakh languages (_-za_ in Ingush, _-zan_ in Chechen and _-c’in_ in Tsova-Tush), and in Armenian (_-at_). Udi (Lezgic), along with the Turkic suffix _-suz_, features the caritive suffix _-nutː_.

In languages lacking a caritive suffix/clitic, caritive semantics is usually expressed by the negative participle of the verb ‘have’. In the case of East Caucasian, the negative participle of an existential verb is used (cf. (@six) above). So far, I have found information about participial constructions expressing caritive semantics in the following languages: Agul, Avar, Bagvalal, Budukh, Chechen, Ingush, Karata, Khwarshi, Kryz, Rutul, Tabasaran, Tindi and Tsakhur. However, I assume that this strategy is available in other languages too.

On [Map 1](027_caritive_attr_map.html) the value `not having participle` indicates situations where a participial construction was mentioned in the sources, but a dedicated suffix/clitic was not. The value `dedicated suffix/clitic (~ less)` indicates situations where a dedicated suffix/clitic is described in the available sources, while the participial construction is or is not mentioned (these two scenarios are merged on the map).


### Adverbial caritive phrases

Within the class of adverbial caritive phrases there are three competing strategies for expressing caritive meanings: a dedicated suffix/clitic, a participle which can be roughly translated as ‘not having (something)’, and an adpositional construction.

Overall, the number of adverbial caritive suffixes/clitics is smaller than that of attributive ones. The most widespread suffix/clitic is again the Turkic suffix _-sız/-siz/-suz/-süz_, which can derive adverbial caritive phrases in Khinalug, at least some Lezgic languages (Budukh, Kryz, Rutul, Tabasaran, Udi), and some Turkic languages (Azerbaijani, Kumyk). Other adverbial caritive markers in the dataset are the Tsova-Tush suffix _-c’i(n)_, the Avar suffix _-q’_ and the Andi marker _ti(-gu)_ with an unclear morphological status. It is important to mention that the Avar suffix _-q’_ is described in the literature as a marker denoting not only the complete absence but also “shortage of some quality of a noun” [@alekseevetal2012: 130], and probably the same applies to its attributive form _-q’-a-b_ discussed above. The Andi marker _ti(-gu)_ (_-gu_ is an adverbializer) is interesting because it can have not just noun stems but entire noun phrases in its scope, as noted in [@maisakpanovadraft]: “in particular, it can follow a noun in the absolutive plural form, or a noun which already hosts another clitic”, cf. (@seven)-(@eight). 

(@seven) Andi, Andi dialect [@maisakpanovadraft]
```{r}
gloss_example("išːi-w-ul=gu saldat-ol **ti-gu**",
              "{[}we.EXCL-M(GEN)-PL=INT soldier-PL{]} **CAR-ADVZ**",
              "without our soldiers")
```

(@eight) Andi, Andi dialect [@maisakpanovadraft]
```{r}
gloss_example("soru hek’ʷa=lo **ti-gu**",
              "{[}together man{]}=ADD CAR-ADVZ",
              "without a single fellow man")
```

In most languages, caritive semantics in adverbial phrases is expressed by means of converbial forms of ‘have’-verbs. At present, the dataset contains information on converbial forms of existential verbs functioning as caritive markers in 25 East Caucasian languages: Agul, Akhvakh, Avar, Bagvalal, Bezhta, Botlikh, Budukh, Chamalal, Chechen, Standard Dargwa, Sanzhi Dargwa, Mehweb Dargwa, Godoberi, Hinuq, Ingush, Karata, Khinalug, Khwarshi, Lak, Lezgian, Rutul, Tabasaran, Tindi, Tsakhur and Tsez. Similar to the situation with participial constructions, it seems that the absence of this type of caritive markers in some other languages can be explained by a lack of information in the literature.

In Armenian the standard way of expressing the caritive meaning in adverbial phrases is to use the preposition _ar°anc’_ (@nine), which apparently represents a separate (third) type of caritive markers attested in the area.

(@nine) Armenian [@dum-tragut2009: 248]
```{r, results='asis'}
gloss_example("et’e et.nayel-u lin-em č’-em kar-oɬ as-el t’e Hayastan-n aysōr ur klin-er **aṙanc’** Sp’yuṙk’-i",
              "CONJ look.back-PTCP.FUT be-SUBJ.FUT.1.SG NEG-be.PRS.1SG can-PRS.PTCP say-INF CONJ Armenia.NOM-the today where be-PST.COND.3.SG **PREP** Diaspora-DAT",
              "If I am about to look back, I cannot say where Armenia would be without the Diaspora.", line_length =90)
```

The principles of encoding the languages for [Map 2](028_caritive_adv_map.html) are the same as for [Map 1](027_caritive_attr_map.html#m2). While the value `not having converb` means that, according to my data, a language has only the converbial strategy, the value `dedicated suffix/clitic (~ lessly)` means that a language uses only a dedicated suffix/clitic or it uses both a dedicated suffix/clitic and a converbial constriction.

### Additional remarks

When compiling the dataset, I tried to follow the classification described above as much as possible. However, there are examples which represent intermediate cases and thus require some additional remarks. 

First, in some languages one of the non-finite verbal forms apparently became the default caritive marker occurring with both attributive and adverbial functions, cf. the participial form _akːʷar_ modifying the predicate in (@ten). In such cases, the language was marked as having both participial and converbial caritive strategies regardless of the morphological make-up of the form.

(@ten)	Sanzhi Dargwa [@forker2020dargwa: 270]
```{r, results='asis'}
gloss_example("li<b>il durħ-ne wec’al durħuˁ xːunul **akːʷ-ar** ha-b-iq’-un-te=de c’il",
              "all<HPL> boy-PL ten boy woman **COP.NEG-PRS** up-HPL-bring.up-PRET-DD.PL=PST then",
              "All children, ten children (he) rose without his wife then.")
```

Second, in many East Caucasian languages the participial and converbial caritive markers are obviously in the process of grammaticalization to postpositions or even further to nominal suffixes. For example, the Sanzhi Dargwa marker _akːʷ-ar_ (@ten), which is historically a non-finite verbal form, is already called ‘postposition’ in the grammatical description [@forker2020dargwa]. In dictionaries we often find examples where the participial and converbial caritive markers are written together with the noun, which suggests their suffixal status, cf. (@eleven).

(@eleven)	Standard Dargwa [@isaev2001: 87]
```{r}
gloss_example("muc’ur-**agar** uqna",
              "beard-**COP.PTCP.NEG** old_man",
              "beardless old man")
```

However, since the degree of grammaticalization of such markers is very hard to estimate, in the dataset all of them are encoded as `not having participle` / `not having converb` based on their (in most cases, obvious) historical origin.


## Distribution

The most interesting areal pattern observed on the maps is the spread of the Turkic caritive suffix _-sız/-siz/-suz/-süz_ to Lezgic languages. Note that Archi, which belongs to the Lezgic branch but is spoken in a different region, lacks this type of suffix, which suggests the contact hypothesis. Also, it can be suggested that dedicated caritive suffixes/clitics (of non-Turkic origin) are more typical for the northern branches (Nakh, Avar-Andic and Tsezic) than for the central branches (Dargwa and Lak) of the East Caucasian family. Nevertheless, even in northern branches the dedicated caritive suffixes/clitics apparently become marginalized and are often replaced by participal/converbial constructions.

## List of glosses {-}

```{r}
make_gloss_list(readxl::read_xlsx("../glosses.xlsx"))
```

## References {-}
