
## Introduction
This chapter is dedicated to presence or absence of the numeral marker in ordinal numerals across the East Caucasian family.

## Results
Ordinals express the index of an object in a row of other objects, e.g. English ‘fourth’. Ordinals have their own suffixes, but in some East Caucasian languages these suffixes also contain numeral markers:

(@) Karata proper [@magomedbekova1971: 96]
```{r, echo=FALSE}
gloss_example(transliteration = "ɬab-**da-ƛo-b**",
              glosses = "three-**NUM-ORD-CM**",
              free_translation = "third"
)
```

In others, the numeral marker is absent:

(@) Tindi [@magomedova2012: 145]
```{r, echo=FALSE}
gloss_example(transliteration = "ɬabi-liƛa-b kalasi",
              glosses = "three-ORD-CM class",
              free_translation = "third class"
)
```

In East Caucasian languages, ordinal markers commonly originate in past participles from the verb ‘say’ with different aspectual features, as shown by [@nasledskovanetkachev2021; @martzloff2017: 156]. This feature is not mapped in WALS, though it includes  a map on expression of the first three natural numerals in the ordinal series [@stolzveselinova2013].

## Distribution
In Dargic languages ordinals usually do not have a numeral marker, while in Lezgic and Tsezic languages they usually do. Apart from that, this feature is rather inconsistent across branches.

## List of glosses
`r add_gloss("CM")` — class marker; `r add_gloss("ORD")` — ordinal

## References
